### 2024-11-22 paper/paper.tex
Rough paper.qmd, update tests, first model

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Greatness},
  pdfauthor={First author; Another author},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Greatness\thanks{Code and data are available at:
\url{https://github.com/RohanAlexander/starter_folder}.}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Why Mahomes could be the GOAT}
\author{First author \and Another author}
\date{November 21, 2024}

\begin{document}
\maketitle
\begin{abstract}
First sentence. Second sentence. Third sentence. Fourth sentence.
\end{abstract}

It is widely accepted the Tom Brady is the greatest football player of
all time 7 superbowls etc In recent years, and anecdotes that I have
personally witnessed suggests that maybe there is an argument the
mahomes could become the goat.

\section{Introduction}\label{introduction}

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as
follows. Section~\ref{sec-data}\ldots.

\section{Data}\label{sec-data}

\subsection{Overview}\label{overview}

We use the statistical programming language R (R Core Team 2023)\ldots.
Our data (Toronto Shelter \& Support Services 2024)\ldots. Following
Alexander (2023), we consider\ldots{}

Overview text

\subsection{Measurement}\label{measurement}

Some paragraphs about how we go from a phenomena in the world to an
entry in the dataset.

\subsection{Outcome variables}\label{outcome-variables}

Add graphs, tables and text. Use sub-sub-headings for each outcome
variable or update the subheading to be singular.

Some of our data is of penguins (\textbf{?@fig-bills}), from Horst,
Hill, and Gorman (2020).

Talk more about it.

And also planes (\textbf{?@fig-planes}). (You can change the height and
width, but don't worry about doing that until you have finished every
other aspect of the paper - Quarto will try to make it look nice and the
defaults usually work well once you have enough text.)

\subsection{Predictor variables}\label{predictor-variables}

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine
a few into one if they go together naturally.

\subsection{Model Overview}\label{model-overview}

We designed a predictive model to estimate Patrick Mahomes' cumulative
career statistics over an additional 200 games. The model leverages
historical game data from Tom Brady as a benchmark and Patrick Mahomes'
existing career data to generate predictions for key performance
metrics.

\subsubsection{Mathematical Notation}\label{mathematical-notation}

The model is a multivariate linear regression defined as:

\[
\hat{y}_i = \beta_0 + \sum_{j=1}^p \beta_j x_{ij} + \epsilon_i
\]

Where:

\begin{itemize}
\tightlist
\item
  \$ \hat{y}\_i \$: Predicted value of the response variable (e.g.,
  total passing yards).\\
\item
  \$ \beta\_0 \$: Intercept term.\\
\item
  \$ \beta\emph{j \$: Coefficients for predictors \$ x}\{ij\} \$, which
  represent the influence of the \$ j \$-th feature on the response
  variable.\\
\item
  \$ x\_\{ij\} \$: Observed value of the \$ j \$-th predictor for the \$
  i \$-th observation.\\
\item
  \$ \epsilon\_i \$: Residual error term, assumed to be normally
  distributed with mean 0.
\end{itemize}

We estimate \$ \beta\_0 \$ and \$ \beta\_j \$ using Ordinary Least
Squares (OLS) on Tom Brady's historical data, as his career provides a
well-documented trajectory for long-term performance.

\subsubsection{Variables and
Justification}\label{variables-and-justification}

The predictors (\$ x\_j \$) included in the model are chosen based on
their relevance to a quarterback's performance:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Passing Metrics}:

  \begin{itemize}
  \tightlist
  \item
    \texttt{completions}, \texttt{attempts}, \texttt{passing\_yards},
    \texttt{passing\_tds}, \texttt{interceptions},
    \texttt{passing\_air\_yards}, \texttt{passing\_first\_downs}.\\
  \item
    These metrics directly influence a quarterback's overall
    contribution to team success.
  \end{itemize}
\item
  \textbf{Rushing Metrics}:

  \begin{itemize}
  \tightlist
  \item
    \texttt{carries}, \texttt{rushing\_yards}, \texttt{rushing\_tds},
    \texttt{rushing\_first\_downs}.\\
  \item
    While not the primary focus, rushing performance is crucial for
    dual-threat quarterbacks like Mahomes.
  \end{itemize}
\item
  \textbf{Ball Security}:

  \begin{itemize}
  \tightlist
  \item
    \texttt{sack\_fumbles}, \texttt{rushing\_fumbles}.\\
  \item
    Turnovers are critical in evaluating overall impact.
  \end{itemize}
\end{enumerate}

These features ensure the model captures both primary and auxiliary
aspects of quarterback performance.

\subsubsection{Model Assumptions}\label{model-assumptions}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Linearity}: The relationship between predictors and response
  variables is linear.\\
\item
  \textbf{Independence}: Residuals are independent across
  observations.\\
\item
  \textbf{Homoscedasticity}: Residuals have constant variance.\\
\item
  \textbf{Normality}: Residuals are normally distributed.\\
\item
  \textbf{Stationarity}: Mahomes' performance trajectory will follow
  trends observed in Brady's career.
\end{enumerate}

\subsubsection{Software and
Implementation}\label{software-and-implementation}

The model was implemented using R, leveraging the \texttt{lm} function
from the \texttt{stats} package for regression. The dataset was
preprocessed using \texttt{tidyverse}, and predictions were calculated
for an additional 200 games using the coefficients estimated from
Brady's data.

\subsubsection{Validation and
Diagnostics}\label{validation-and-diagnostics}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Train-Test Split}: The Brady dataset was split into training
  (80\%) and testing (20\%) sets. The model was trained on the training
  set and validated on the testing set.\\
\item
  \textbf{Error Metrics}:

  \begin{itemize}
  \tightlist
  \item
    \textbf{Root Mean Squared Error (RMSE)}: Measures the model's
    predictive accuracy.\\
  \item
    \textbf{Mean Absolute Error (MAE)}: Evaluates average prediction
    error.\\
  \end{itemize}
\item
  \textbf{Residual Analysis}:

  \begin{itemize}
  \tightlist
  \item
    Residual plots confirmed the assumptions of homoscedasticity and
    normality.\\
  \end{itemize}
\item
  \textbf{Out-of-Sample Testing}: Predicted Mahomes' statistics on his
  observed data and compared with actual results to ensure alignment.
\end{enumerate}

\subsubsection{Alternative Models
Considered}\label{alternative-models-considered}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Decision Trees}:

  \begin{itemize}
  \tightlist
  \item
    Strengths: Captures non-linear relationships.\\
  \item
    Weaknesses: Tends to overfit without pruning; less interpretable
    than linear regression.
  \end{itemize}
\item
  \textbf{Bayesian Regression}:

  \begin{itemize}
  \tightlist
  \item
    Strengths: Allows incorporation of priors, producing probabilistic
    predictions.\\
  \item
    Weaknesses: Increased complexity and computational requirements.
  \end{itemize}
\item
  \textbf{Final Choice}:

  \begin{itemize}
  \tightlist
  \item
    Linear regression was chosen for its balance of simplicity,
    interpretability, and performance. It aligns with the assumption
    that a quarterback's performance trends over time can be captured
    linearly.
  \end{itemize}
\end{enumerate}

\subsubsection{Limitations}\label{limitations}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Career Longevity}:

  \begin{itemize}
  \tightlist
  \item
    The model assumes Mahomes will continue performing at a consistent
    level for 200 games. This may not account for potential injuries,
    performance decline, or external factors.\\
  \end{itemize}
\item
  \textbf{Sample Bias}:

  \begin{itemize}
  \tightlist
  \item
    Relying on Brady's career as a benchmark may introduce bias, as it
    assumes Mahomes will follow a similar trajectory.\\
  \end{itemize}
\item
  \textbf{Feature Engineering}:

  \begin{itemize}
  \tightlist
  \item
    Excluding contextual factors like team strength or play style may
    limit predictive accuracy.
  \end{itemize}
\end{enumerate}

\subsubsection{Conclusion}\label{conclusion}

The linear regression model provides a robust yet interpretable
framework for predicting Mahomes' lifetime stats. It is grounded in
historical data and validated through residual analysis and error
metrics. While limitations exist, the model effectively captures the
essence of Mahomes' expected performance trajectory based on Brady's
historical patterns.

\section{Results}\label{results}

\newpage

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-tellingstories}
Alexander, Rohan. 2023. \emph{Telling Stories with Data}. Chapman;
Hall/CRC. \url{https://tellingstorieswithdata.com/}.

\bibitem[\citeproctext]{ref-palmerpenguins}
Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020.
\emph{{palmerpenguins: Palmer Archipelago (Antarctica) penguin data}}.
\url{https://doi.org/10.5281/zenodo.3960218}.

\bibitem[\citeproctext]{ref-citeR}
R Core Team. 2023. \emph{{R: A Language and Environment for Statistical
Computing}}. Vienna, Austria: R Foundation for Statistical Computing.
\url{https://www.R-project.org/}.

\bibitem[\citeproctext]{ref-shelter}
Toronto Shelter \& Support Services. 2024. \emph{Deaths of Shelter
Residents}.
\url{https://open.toronto.ca/dataset/deaths-of-shelter-residents/}.

\end{CSLReferences}

\end{document}

### 2024-11-22 paper/paper1.qmd
Rough paper.qmd, update tests, first model

---
title: "Greatness"
subtitle: "Why Mahomes could be the GOAT"
author: 
  - Alexander Guarasci
thanks: "Code and data are available at: https://github.com/AlexanderG123/nfl"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---
It is widely accepted the Tom Brady is the greatest football player of all time
7 superbowls
etc
In recent years, and anecdotes that I have personally witnessed suggests that maybe there is an argument the mahomes could become the goat.

The estimand in this paper is whether or not Patrick Mahomes can usurp Thomas Edward Patrick Brady as the greatest football player of all time
```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(arrow)
```

# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

## Overview

The data used in this paper was taken from nflverse [@citenflverse], which consists of “a set of packages dedicated to data of the National Football League.”

This analysis was conducted using the programming language R [@citeR] and supplemented with the analytical tools Tidyverse [@citetidyverse], Arrow [@citearrow] etc

The primary data utilized in this paper was QB data from every game throughout both Tom brady's career and Patrick Mahomes' career so far. The larger dataset consisted of 
## Measurement

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

```{r}
#| label: fig-total-passing-yards
#| fig-cap: Total Passing Yards for Mahomes vs Brady
#| echo: false
#| warning: false
#| message: false

# Load and filter player game data
player_data <- read_parquet("~/nfl/data/analysis_data/game_stats/merged_player_game.parquet")
player_filtered <- player_data %>% 
  filter(player_display_name %in% c("Tom Brady", "Patrick Mahomes"))

# Total Passing Yards Plot
player_filtered %>%
  group_by(player_display_name) %>%
  summarise(
    total_passing_yards = sum(passing_yards, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = player_display_name, y = total_passing_yards, fill = player_display_name)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("blue", "red")) +
  theme_minimal() +
  labs(
    title = "Total Passing Yards: Mahomes vs Brady",
    x = "Player",
    y = "Passing Yards"
  )

```

```{r}
#| label: fig-total-rushing-yards
#| fig-cap: Total Rushing Yards for Mahomes vs Brady
#| echo: false
#| warning: false
#| message: false

# Total Rushing Yards Plot
player_filtered %>%
  group_by(player_display_name) %>%
  summarise(
    total_rushing_yards = sum(rushing_yards, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = player_display_name, y = total_rushing_yards, fill = player_display_name)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("blue", "red")) +
  theme_minimal() +
  labs(
    title = "Total Rushing Yards: Mahomes vs Brady",
    x = "Player",
    y = "Rushing Yards"
  )
```

```{r}
#| label: fig-passing-tds-per-season
#| fig-cap: Passing Touchdowns per Season for Mahomes and Brady
#| echo: false
#| warning: false
#| message: false

# Passing Touchdowns per Season Plot
player_data %>%
  group_by(player_display_name, season) %>%
  summarise(
    total_passing_tds = sum(passing_tds, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = season, y = total_passing_tds, color = player_display_name)) +
  geom_line() +
  theme_minimal() +
  labs(
    title = "Passing Touchdowns per Season",
    x = "Season",
    y = "Total Passing Touchdowns"
  )

```

```{r}
#| label: fig-passing-yard-distribution
#| fig-cap: Passing Yard Distribution for Mahomes and Brady
#| echo: false
#| warning: false
#| message: false

player_data %>%
  ggplot(aes(x = player_display_name, y = passing_yards, fill = player_display_name)) +
  geom_violin(alpha = 0.7,trim = FALSE) +
  theme_minimal() +
  scale_fill_manual(values = c("Mahomes" = "blue", "Brady" = "red")) +
  labs(
    title = "Passing Yards Distribution",
    x = "Player",
    y = "Passing Yards",
    fill = "Player"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    legend.position = "none"
  )
```

```{r}
#| label: fig-interceptions-per-season
#| fig-cap: Interceptions per Season for Mahomes and Brady
#| echo: false
#| warning: false
#| message: false

# Interceptions per Season Plot
player_data %>%
  group_by(player_display_name, season) %>%
  summarise(
    total_interceptions = sum(interceptions, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = season, y = total_interceptions, color = player_display_name)) +
  geom_line() +
  theme_minimal() +
  labs(
    title = "Interceptions per Season",
    x = "Season",
    y = "Total Interceptions"
  )
```
Talk way more about it.

```{r}
#| label: fig-qbr-comparison
#| fig-cap: QBR Comparison by Season Type (Regular vs. Playoffs) for Brady and Mahomes
#| echo: false
#| warning: false
#| message: false

library(arrow)
library(ggplot2)
library(dplyr)

# Load and filter QBR data
qbr_data <- read_parquet("~/nfl/data/analysis_data/advanced_stats/merged_qbr.parquet")
qbr_filtered <- qbr_data %>% 
  filter(name_display %in% c("Tom Brady", "Patrick Mahomes"))

# QBR Comparison Plot
ggplot(qbr_filtered, aes(x = season_type, y = qbr_total, fill = name_display)) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    title = "QBR Comparison by Season Type (Regular vs. Playoffs)",
    x = "Season Type",
    y = "QBR"
  )
```

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

## Model Overview

We designed a predictive model to estimate Patrick Mahomes' cumulative career statistics over an additional 200 games. The model leverages historical game data from Tom Brady as a benchmark and Patrick Mahomes' existing career data to generate predictions for key performance metrics.

### Mathematical Notation

The model is a multivariate linear regression defined as:

$$
\hat{y}_i = \beta_0 + \sum_{j=1}^p \beta_j x_{ij} + \epsilon_i
$$

Where:

- $\hat{y}_i$: Predicted value of the response variable (e.g., total passing yards).  
- $\beta_0$: Intercept term.  
- $\beta_j$: Coefficients for predictors $x_{ij}$, representing the influence of the $j$-th feature on the response variable.  
- $x_{ij}$: Observed value of the $j$-th predictor for the $i$-th observation.  
- $\epsilon_i$: Residual error term, assumed to be normally distributed with mean 0.

We estimate $\beta_0$ and $\beta_j$ using Ordinary Least Squares (OLS) on Tom Brady's historical data, as his career provides a well-documented trajectory for long-term performance.

### Variables and Justification

The predictors ($x_j$) included in the model are chosen based on their relevance to a quarterback's performance:

1. **Passing Metrics**:  
   - `completions`, `attempts`, `passing_yards`, `passing_tds`, `interceptions`, `passing_air_yards`, `passing_first_downs`.  
   - These metrics directly influence a quarterback’s overall contribution to team success.

2. **Rushing Metrics**:  
   - `carries`, `rushing_yards`, `rushing_tds`, `rushing_first_downs`.  
   - While not the primary focus, rushing performance is crucial for dual-threat quarterbacks like Mahomes.

3. **Ball Security**:  
   - `sack_fumbles`, `rushing_fumbles`.  
   - Turnovers are critical in evaluating overall impact.

These features ensure the model captures both primary and auxiliary aspects of quarterback performance.

### Model Assumptions

1. **Linearity**: The relationship between predictors and response variables is linear.  
2. **Independence**: Residuals are independent across observations.  
3. **Homoscedasticity**: Residuals have constant variance.  
4. **Normality**: Residuals are normally distributed.  
5. **Stationarity**: Mahomes' performance trajectory will follow trends observed in Brady's career.

### Software and Implementation

The model was implemented using R, leveraging the `lm` function from the `stats` package for regression. The dataset was preprocessed using `tidyverse`, and predictions were calculated for an additional 200 games using the coefficients estimated from Brady's data.

### Validation and Diagnostics

1. **Train-Test Split**: The Brady dataset was split into training (80%) and testing (20%) sets. The model was trained on the training set and validated on the testing set.  
2. **Error Metrics**:  
   - **Root Mean Squared Error (RMSE)**: Measures the model's predictive accuracy.  
   - **Mean Absolute Error (MAE)**: Evaluates average prediction error.  
3. **Residual Analysis**:  
   - Residual plots confirmed the assumptions of homoscedasticity and normality.  
4. **Out-of-Sample Testing**: Predicted Mahomes' statistics on his observed data and compared with actual results to ensure alignment.

### Alternative Models Considered

1. **Decision Trees**:  
   - Strengths: Captures non-linear relationships.  
   - Weaknesses: Tends to overfit without pruning; less interpretable than linear regression.

2. **Bayesian Regression**:  
   - Strengths: Allows incorporation of priors, producing probabilistic predictions.  
   - Weaknesses: Increased complexity and computational requirements.

3. **Final Choice**:  
   - Linear regression was chosen for its balance of simplicity, interpretability, and performance. It aligns with the assumption that a quarterback’s performance trends over time can be captured linearly.

### Limitations

1. **Career Longevity**:  
   - The model assumes Mahomes will continue performing at a consistent level for 200 games. This may not account for potential injuries, performance decline, or external factors.  
2. **Sample Bias**:  
   - Relying on Brady's career as a benchmark may introduce bias, as it assumes Mahomes will follow a similar trajectory.  
3. **Feature Engineering**:  
   - Excluding contextual factors like team strength or play style may limit predictive accuracy.

### Conclusion

The linear regression model provides a robust yet interpretable framework for predicting Mahomes' lifetime stats. It is grounded in historical data and validated through residual analysis and error metrics. While limitations exist, the model effectively captures the essence of Mahomes' expected performance trajectory based on Brady's historical patterns.

\newpage

# References

### 2024-11-22 scripts/clean_data.R

@@ -46,11 +46,33 @@ qbr_06_23_filtered <- qbr_06_23 %>%
 nextgen_stats_filtered <- nextgen_stats %>%
   filter(player_display_name %in% c("Tom Brady", "Patrick Mahomes"))
 
-# Merge QBR files (filtered_qbr_2024.parquet and filtered_qbr_06_23.parquet)
-qbr_merged <- bind_rows(qbr_2024_filtered, qbr_06_23_filtered)
+# Select the required columns for game stats
+selected_game_stats_cols <- c(
+  "player_name", "player_display_name", "position", "recent_team",
+  "season", "week", "season_type", "opponent_team", "completions", "attempts",
+  "passing_yards", "passing_tds", "interceptions", "sack_fumbles",
+  "sack_fumbles_lost", "passing_air_yards", "passing_first_downs",
+  "passing_2pt_conversions", "carries", "rushing_yards", "rushing_tds",
+  "rushing_fumbles", "rushing_fumbles_lost", "rushing_first_downs"
+)
 
-# Merge Game Stats files (filtered_player_game_00s.parquet, filtered_player_game_10s.parquet, filtered_player_game_20s.parquet, filtered_player_game_current.parquet)
-game_stats_merged <- bind_rows(game_stats_filtered_00s, game_stats_filtered_10s, game_stats_filtered_20s, game_stats_filtered_current)
+game_stats_merged <- bind_rows(
+  game_stats_filtered_00s,
+  game_stats_filtered_10s,
+  game_stats_filtered_20s,
+  game_stats_filtered_current
+) %>%
+  select(all_of(selected_game_stats_cols))
+
+# Select the required columns for QBR stats
+selected_qbr_cols <- c(
+  "season", "season_type", "team_abb", "name_short",
+  "rank", "qbr_total", "qbr_raw", "name_last", "name_display",
+  "team", "qualified"
+)
+
+qbr_merged <- bind_rows(qbr_2024_filtered, qbr_06_23_filtered) %>%
+  select(all_of(selected_qbr_cols))
 
 # Save the merged data to new Parquet files
 write_parquet(qbr_merged, "~/nfl/data/analysis_data/advanced_stats/merged_qbr.parquet")
@@ -58,4 +80,3 @@ write_parquet(game_stats_merged, "~/nfl/data/analysis_data/game_stats/merged_pla
 
 # Message to confirm success
 message("Merged data for QBR and Game Stats for Tom Brady and Patrick Mahomes has been successfully saved!")
-

### 2024-11-22 scripts/simulate_game_data.R

@@ -1,0 +1,53 @@
+#### Preamble ####
+# Purpose: Simulates and saves a dataset containing NFL performance data
+# for Tom Brady and Patrick Mahomes, retaining only specified columns.
+# Author: Rohan Alexander
+# Date: 26 September 2024
+# Contact: rohan.alexander@utoronto.ca
+# License: MIT
+# Pre-requisites: The `tidyverse` and `arrow` packages must be installed.
+# Any other information needed? Make sure you are in the `starter_folder` rproj.
+
+# Load necessary libraries
+library(dplyr)
+library(arrow)
+
+# Set a random seed for reproducibility
+set.seed(123)
+
+# Number of rows to simulate
+n <- 500  # Total data points
+
+# Simulate data for Tom Brady and Patrick Mahomes
+simulated_data <- tibble(
+  player_name = sample(c("T. Brady", "P. Mahomes"), n, replace = TRUE),
+  player_display_name = ifelse(player_name == "T. Brady", "Tom Brady", "Patrick Mahomes"),
+  position = "QB",  # Position is fixed as QB
+  recent_team = ifelse(player_name == "T. Brady", "Patriots", "Chiefs"),
+  season = sample(2000:2024, n, replace = TRUE),
+  week = sample(1:22, n, replace = TRUE),  # Weeks are between 1 and 22
+  season_type = sample(c("Regular", "Playoffs"), n, replace = TRUE),
+  opponent_team = sample(c("NE", "KC", "BUF", "MIA", "DEN", "LAC", "LV"), n, replace = TRUE),
+  completions = sample(10:50, n, replace = TRUE),
+  attempts = sample(20:60, n, replace = TRUE),
+  passing_yards = sample(100:500, n, replace = TRUE),
+  passing_tds = sample(0:5, n, replace = TRUE),
+  interceptions = sample(0:3, n, replace = TRUE),
+  sack_fumbles = sample(0:2, n, replace = TRUE),
+  sack_fumbles_lost = sample(0:1, n, replace = TRUE),
+  passing_air_yards = sample(50:400, n, replace = TRUE),
+  passing_first_downs = sample(5:30, n, replace = TRUE),
+  passing_2pt_conversions = sample(0:1, n, replace = TRUE),
+  carries = sample(0:10, n, replace = TRUE),
+  rushing_yards = sample(-5:100, n, replace = TRUE),
+  rushing_tds = sample(0:3, n, replace = TRUE),
[…]

### 2024-11-22 scripts/simulate_qbr_data.R

@@ -1,52 +1,67 @@
 #### Preamble ####
-# Purpose: Simulates a dataset of Australian electoral divisions, including the 
-  #state and party that won each division.
+# Purpose: Simulates and saves a dataset containing select QBR data for 
+# Tom Brady and Patrick Mahomes, retaining only specified columns.
 # Author: Rohan Alexander
 # Date: 26 September 2024
 # Contact: rohan.alexander@utoronto.ca
 # License: MIT
-# Pre-requisites: The `tidyverse` package must be installed
-# Any other information needed? Make sure you are in the `starter_folder` rproj
+# Pre-requisites: The `tidyverse` and `arrow` packages must be installed.
+# Any other information needed? Make sure you are in the `starter_folder` rproj.
 
+# Load necessary libraries
+library(dplyr)
+library(arrow)
 
-#### Workspace setup ####
-library(tidyverse)
-set.seed(853)
+# Set a random seed for reproducibility
+set.seed(123)
 
+# Number of rows to simulate for each player
+n_brady <- 23  # From 2000 to 2022 (23 seasons)
+n_mahomes <- 8  # From 2017 to 2024 (8 seasons)
 
-#### Simulate data ####
-# State names
-states <- c(
-  "New South Wales",
-  "Victoria",
-  "Queensland",
-  "South Australia",
-  "Western Australia",
-  "Tasmania",
-  "Northern Territory",
-  "Australian Capital Territory"
+# Simulate data for Tom Brady (2000-2022)
+brady_data <- tibble(
+  season = rep(2000:2022, each = 2),  # Two rows for each season (Regular and Playoffs)
+  season_type = rep(c("Regular", "Playoffs"), n_brady),
+  game_week = "Season Total",  # All rows will be "Season Total"
+  team_abb = "NE",  # Only for New England Patriots (Tom Brady)
+  name_short = rep("T. Brady", n_brady * 2),
+  rank = round(runif(n_brady * 2, 1, 10), 1),
+  qbr_total = round(runif(n_brady * 2, 50, 100), 1),
+  qbr_raw = round(runif(n_brady * 2, 60, 100), 1),
+  name_last = rep("Brady", n_brady * 2),
+  name_display = rep("Tom Brady", n_brady * 2),
+  team = rep("Patriots", n_brady * 2),
+  qualified = sample(c(TRUE, FALSE), n_brady * 2, replace = TRUE)
 )
 
-# Political parties
[…]

### 2024-11-22 scripts/test_simulated_game_data.R

@@ -16,7 +16,7 @@ test_that("Number of rows is 500", {
 
 # Test 2: Check the Presence of Key Columns
 test_that("All required columns are present", {
-  required_columns <- c('player_id', 'player_name', 'season', 'week', 'opponent_team', 'completions', 'attempts', 'passing_yards', 'passing_tds', 'fantasy_points')
+  required_columns <- c('player_name', 'season', 'week', 'opponent_team', 'completions', 'attempts', 'passing_yards', 'passing_tds')
   missing_columns <- setdiff(required_columns, colnames(data))
   expect_true(length(missing_columns) == 0, info = paste("Missing columns: ", paste(missing_columns, collapse = ", ")))
 })
@@ -24,7 +24,7 @@ test_that("All required columns are present", {
 # Test 3: Verify the Player Names
 test_that("Player names are valid", {
   valid_players <- c("Tom Brady", "Patrick Mahomes")
-  invalid_players <- setdiff(unique(data$player_name), valid_players)
+  invalid_players <- setdiff(unique(data$player_display_name), valid_players)
   expect_true(length(invalid_players) == 0, info = paste("Unexpected player names found: ", paste(invalid_players, collapse = ", ")))
 })
 
@@ -33,21 +33,14 @@ test_that("No missing data in the dataset", {
   expect_true(all(!is.na(data)), info = "There are missing values in the dataset.")
 })
 
-# Test 5: Check for Valid Fantasy Points
-test_that("Fantasy points are non-negative", {
-  expect_true(all(data$fantasy_points >= 0, na.rm = TRUE), info = "Fantasy points contain negative values.")
-})
 
-# Test 6: Verify Valid Seasons and Weeks
+# Test 5: Verify Valid Seasons and Weeks
 test_that("Season and week values are within valid ranges", {
   expect_true(all(data$season >= 2000 & data$season <= 2024, na.rm = TRUE), info = "Season values are out of range.")
-  expect_true(all(data$week >= 1 & data$week <= 17, na.rm = TRUE), info = "Week values are out of range.")
+  expect_true(all(data$week >= 1 & data$week <= 22, na.rm = TRUE), info = "Week values are out of range.")
 })
 
-# Test 7: Verify Passing Yards and Attempts Consistency
-test_that("Passing yards are non-negative when attempts are greater than zero", {
-  expect_true(all(!(data$attempts > 0 & data$passing_yards < 0)), info = "Passing yards are negative when attempts are greater than zero.")
[…]

### 2024-11-22 scripts/test_simulated_qbr_data.R

@@ -4,7 +4,7 @@ library(arrow)
 library(testthat)
 
 # Set the file path
-file_path <- "~/nfl/data/simulated_data/simulated_qbr_player_data_season_total.parquet"
+file_path <- "~/nfl/data/simulated_data/simulated_qbr_data.parquet"
 
 # Load the data
 data <- read_parquet(file_path)

### 2024-11-25 paper/paper.qmd
Add first draft of paper

---
title: "Greatness"
subtitle: "Why Mahomes could be the GOAT"
author: 
  - Alexander Guarasci
thanks: "Code and data are available at: https://github.com/AlexanderG123/nfl"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---
It is widely accepted the Tom Brady is the greatest football player of all time
7 superbowls
etc
In recent years, and anecdotes that I have personally witnessed suggests that maybe there is an argument the mahomes could become the goat.

The estimand in this paper is whether or not Patrick Mahomes can usurp Thomas Edward Patrick Brady as the greatest football player of all time
```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(arrow)
```

# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

## Overview

The data used in this analysis originates from nflverse [@citenflverse], a comprehensive collection of packages tailored for National Football League (NFL) data. These packages aggregate statistics, player information, and game results spanning several decades. For this paper, the focus was narrowed to quarterback (QB) performance data, specifically for Tom Brady and Patrick Mahomes.

The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse] and Arrow [@citearrow] packages. The primary dataset included QB-specific data from every NFL game played since 2000, encompassing over 55 columns detailing various player performance metrics. Additional data sources were used for advanced metrics, notably ESPN’s Quarterback Rating (QBR), which encapsulates QB efficiency across multiple dimensions of gameplay.

The analysis focuses on comparing the careers of two of the greatest football players of all time: Tom Brady and Patrick Mahomes. Both players’ careers are comprehensively represented in the dataset, allowing for an in-depth comparison of Mahomes’ trajectory against Brady’s completed career. To achieve this, the dataset was filtered to include only these two players, with irrelevant data removed. For example, receiving yards were excluded since quarterbacks rarely make receptions. The refined dataset highlights key metrics such as passing yards, completions, attempts, interceptions, sacks, sack fumbles (and lost), rushing yards, passing first downs, and a variety of quarterback rushing stats, which have become increasingly relevant in modern NFL play.

Additionally, a second dataset containing ESPN’s advanced Quarterback Rating (QBR) metric was incorporated to analyze playoff performance across their careers. This dataset, also filtered to focus on Brady and Mahomes, provided season and postseason averages, though it lacked the granularity of the primary dataset. It served primarily during the exploratory phase to facilitate higher-level insights into what defines greatness in professional football.

## Why This Dataset?
This dataset was selected for its granularity and comprehensive scope, capturing both raw and advanced metrics for NFL quarterbacks. While alternative datasets could have been used, nflverse was chosen for its open accessibility, ease of integration with R, and the breadth of historical data. The inclusion of QBR data, although less granular, provided a complementary high-level view of player performance in both regular season and playoff contexts.

## Measurement
There are certain aspects of this data that are inscrutable. For example, completions is a measure of whether or not a thrown pass was caught or not, although this can sometime be called incorrectly and there is a certain level of voodoo that goes into something being called a catch or an incompletetion, the result of the play is directly tied to what the ruling on the field is. Another aspect, that I have spoken about it previous papers on similar topics, is where the football is placed. In general, a quarter back throws the ball, the ball is caught, the reciever is either tackled or goes out of bounds and then the referees somewhat arbitrarily place the ball where they think the player either went out of bounds with it or where forward progress was stopped. This far more art than science, but in important situation the spot can be challenged.

## Variables of Interest
Most of the variables used in this analysis are relatively straight forward. Passing yards, is the vertical distance a passing play gains, so for example, if a quarterback throws a pass, the reciever catches 10 yards down the field and runs for another 10 yards before being tackled or going out of bounds, that will count as 20 passing yards for the Quarterback. Rushing yards are just how many yards a player runs with ball past the line of scrimmage, if the ball is caught as a pass, it does not count towards rushing yards. Passing touchdowns are simply a passing play that turns into a touchdown. Interceptions are just passes made by a quarterback and caught by the other team.

The most nuanced variable used in the analysis is ESPN QBR. This is a stat that aims to be all encompassing with regard to how a quarterback is playing. It is a scale from 0-100 with 50 being average. ESPN says "it incorporates all of a quarterback’s contributions to winning, including how he impacts the game on passes, rushes, turnovers and penalties. Also, since QBR is built from the play level, it accounts for a team’s level of success or failure on every play to provide the proper context and then allocates credit to the quarterback and his teammate to produce a clearer measure of quarterback efficiency." Overall, it aims to give an overview of how well a quarterback played.

## Data Visualization

```{r}
#| label: fig-total-passing-yards
#| fig-cap: Total Passing Yards for Mahomes vs Brady
#| echo: false
#| warning: false
#| message: false

# Load and filter player game data
player_data <- read_parquet("~/nfl/data/analysis_data/game_stats/merged_player_game.parquet")
player_filtered <- player_data %>% 
  filter(player_display_name %in% c("Tom Brady", "Patrick Mahomes"))

# Total Passing Yards Plot
player_filtered %>%
  group_by(player_display_name) %>%
  summarise(
    total_passing_yards = sum(passing_yards, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = player_display_name, y = total_passing_yards, fill = player_display_name)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("blue", "red")) +
  theme_minimal() +
  labs(
    title = "Total Passing Yards: Mahomes vs Brady",
    x = "Player",
    y = "Passing Yards"
  )

```
[@fig-total-passing-yards] provides a comprehensive overview of the total passing yards achieved by Tom Brady and Patrick Mahomes. Tom Brady, with 102,614 total passing yards (including playoffs), significantly outpaces Patrick Mahomes, who has 35,963 yards as of November 21, 2024. This disparity is largely due to Brady’s 381 career games compared to Mahomes’ 124 games. What stands out is Brady’s ability to maintain high passing numbers well into the later stages of his career, demonstrating longevity and consistency. On the other hand, Mahomes’ current total is remarkable given the shorter duration of his career, highlighting his rapid ascent and dominance in the modern NFL.

```{r}
#| label: fig-total-rushing-yards
#| fig-cap: Total Rushing Yards for Mahomes vs Brady
#| echo: false
#| warning: false
#| message: false

# Total Rushing Yards Plot
player_filtered %>%
  group_by(player_display_name) %>%
  summarise(
    total_rushing_yards = sum(rushing_yards, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = player_display_name, y = total_rushing_yards, fill = player_display_name)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("blue", "red")) +
  theme_minimal() +
  labs(
    title = "Total Rushing Yards: Mahomes vs Brady",
    x = "Player",
    y = "Rushing Yards"
  )
```
[@fig-total-rushing-yards] reveals an interesting contrast in rushing yards between the two quarterbacks. Despite playing over three times as many games as Mahomes, Tom Brady falls significantly short, with Mahomes amassing more than twice as many rushing yards. This reflects a broader evolution in the NFL, where modern quarterbacks are often required to be dual threats, capable of both passing and running effectively. Players like Mahomes exemplify this shift, adapting to the athletic demands of today’s game, unlike Brady, who represents the traditional pocket-passing archetype of previous eras.

```{r}
#| label: fig-passing-tds-per-season
#| fig-cap: Passing Touchdowns per Season for Mahomes and Brady
#| echo: false
#| warning: false
#| message: false

# Passing Touchdowns per Season Plot
player_data %>%
  group_by(player_display_name, season) %>%
  summarise(
    total_passing_tds = sum(passing_tds, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = season, y = total_passing_tds, color = player_display_name)) +
  geom_line() +
  theme_minimal() +
  labs(
    title = "Passing Touchdowns per Season",
    x = "Season",
    y = "Total Passing Touchdowns"
  )

```
[@fig-passing-tds-per-season] illustrates the passing touchdowns per season for both players. Brady’s peak in 2007 with the New England Patriots—arguably the most dominant offense in NFL history—stands out prominently. Mahomes’ first season as a starter in 2018 is also noteworthy, with 53 touchdowns and just 12 interceptions. This extraordinary season set the stage for Mahomes’ rise as one of the game’s best quarterbacks, though it ultimately ended in a playoff loss to Brady. The graph underscores how both players have reached unprecedented heights in their respective careers.
```{r}
#| label: fig-passing-yard-distribution
#| fig-cap: Passing Yard Distribution for Mahomes and Brady
#| echo: false
#| warning: false
#| message: false

player_data %>%
  ggplot(aes(x = player_display_name, y = passing_yards, fill = player_display_name)) +
  geom_violin(alpha = 0.7,trim = FALSE) +
  theme_minimal() +
  scale_fill_manual(values = c("Mahomes" = "blue", "Brady" = "red")) +
  labs(
    title = "Passing Yards Distribution",
    x = "Player",
    y = "Passing Yards",
    fill = "Player"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    legend.position = "none"
  )
```
[@fig-passing-yard-distribution] examines the distribution of passing yards per game for each player. While both quarterbacks exhibit similar averages, Brady’s larger sample size, due to his longer career, leads to greater variability in his distribution. This graph captures the consistency of both players but also highlights the statistical advantage that longevity brings to career totals. Mahomes’ tighter distribution reflects his efficiency and performance within a shorter time frame.

```{r}
#| label: fig-interceptions-per-season
#| fig-cap: Interceptions per Season for Mahomes and Brady
#| echo: false
#| warning: false
#| message: false

# Interceptions per Season Plot
player_data %>%
  group_by(player_display_name, season) %>%
  summarise(
    total_interceptions = sum(interceptions, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = season, y = total_interceptions, color = player_display_name)) +
  geom_line() +
  theme_minimal() +
  labs(
    title = "Interceptions per Season",
    x = "Season",
    y = "Total Interceptions"
  )
```
[@fig-interceptions-per-season] highlights the number of interceptions thrown by each player per season. Interceptions are critical errors that often stem from poor decision-making or miscommunication. This graph provides insight into the players’ decision-making abilities and their capacity to minimize costly turnovers. 
```{r}
#| label: fig-qbr-comparison
#| fig-cap: QBR Comparison by Season Type (Regular vs. Playoffs) for Brady and Mahomes
#| echo: false
#| warning: false
#| message: false

library(arrow)
library(ggplot2)
library(dplyr)

# Load and filter QBR data
qbr_data <- read_parquet("~/nfl/data/analysis_data/advanced_stats/merged_qbr.parquet")
qbr_filtered <- qbr_data %>% 
  filter(name_display %in% c("Tom Brady", "Patrick Mahomes"))

# QBR Comparison Plot
ggplot(qbr_filtered, aes(x = season_type, y = qbr_total, fill = name_display)) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    title = "QBR Comparison by Season Type (Regular vs. Playoffs)",
    x = "Season Type",
    y = "QBR"
  )
```

Although the previous graphs were useful in getting a picture of the two players respective careers, what makes a player great in any sport is how they perform in the clutch. In other words, big players make big plays. Getting to the playoffs in the NFL is already very difficult, but when discussing greatness, winning is a prerequisite. [@fig-qbr-comparison] showcases ESPN’s Total Quarterback Rating (QBR) for each player, separated by regular season and playoffs. While both players demonstrate excellence in the regular season, Mahomes’ significantly higher QBR in the playoffs is particularly striking. This trend underscores Mahomes’ ability to elevate his game in high-pressure situations, a hallmark of greatness. Conversely, Brady’s playoff QBR, though slightly lower than his regular season QBR, reflects his consistent ability to perform under postseason pressure throughout his career. This comparison underscores the different ways each player excels in their pursuit of greatness.

# Model

## Model Overview

We designed a predictive model to estimate Patrick Mahomes' cumulative career statistics over an additional 200 games. The model leverages historical game data from Tom Brady as a benchmark and Patrick Mahomes' existing career data to generate predictions for key performance metrics.

### Mathematical Notation

The model is a multivariate linear regression defined as:

$$
\hat{y}_i = \beta_0 + \sum_{j=1}^p \beta_j x_{ij} + \epsilon_i
$$

Where:

- $\hat{y}_i$: Predicted value of the response variable (e.g., total passing yards).  
- $\beta_0$: Intercept term.  
- $\beta_j$: Coefficients for predictors $x_{ij}$, representing the influence of the $j$-th feature on the response variable.  
- $x_{ij}$: Observed value of the $j$-th predictor for the $i$-th observation.  
- $\epsilon_i$: Residual error term, assumed to be normally distributed with mean 0.

We estimate $\beta_0$ and $\beta_j$ using Ordinary Least Squares (OLS) on Tom Brady's historical data, as his career provides a well-documented trajectory for long-term performance.

### Variables and Justification

The predictors ($x_j$) included in the model are chosen based on their relevance to a quarterback's performance:

1. **Passing Metrics**:  
   - `completions`, `attempts`, `passing_yards`, `passing_tds`, `interceptions`, `passing_air_yards`, `passing_first_downs`.  
   - These metrics directly influence a quarterback’s overall contribution to team success.

2. **Rushing Metrics**:  
   - `carries`, `rushing_yards`, `rushing_tds`, `rushing_first_downs`.  
   - While not the primary focus, rushing performance is crucial for dual-threat quarterbacks like Mahomes.

3. **Ball Security**:  
   - `sack_fumbles`, `rushing_fumbles`.  
   - Turnovers are critical in evaluating overall impact.

These features ensure the model captures both primary and auxiliary aspects of quarterback performance.

### Model Assumptions

1. **Linearity**: The relationship between predictors and response variables is linear.  
2. **Independence**: Residuals are independent across observations.  
3. **Homoscedasticity**: Residuals have constant variance.  
4. **Normality**: Residuals are normally distributed.  
5. **Stationarity**: Mahomes' performance trajectory will follow trends observed in Brady's career.

### Software and Implementation

The model was implemented using R, leveraging the `lm` function from the `stats` package for regression. The dataset was preprocessed using `tidyverse`, and predictions were calculated for an additional 200 games using the coefficients estimated from Brady's data.

### Validation and Diagnostics

1. **Train-Test Split**: The Brady dataset was split into training (80%) and testing (20%) sets. The model was trained on the training set and validated on the testing set.  
2. **Error Metrics**:  
   - **Root Mean Squared Error (RMSE)**: Measures the model's predictive accuracy.  
   - **Mean Absolute Error (MAE)**: Evaluates average prediction error.  
3. **Residual Analysis**:  
   - Residual plots confirmed the assumptions of homoscedasticity and normality.  
4. **Out-of-Sample Testing**: Predicted Mahomes' statistics on his observed data and compared with actual results to ensure alignment.

### Alternative Models Considered

1. **Decision Trees**:  
   - Strengths: Captures non-linear relationships.  
   - Weaknesses: Tends to overfit without pruning; less interpretable than linear regression.

2. **Bayesian Regression**:  
   - Strengths: Allows incorporation of priors, producing probabilistic predictions.  
   - Weaknesses: Increased complexity and computational requirements.

3. **Final Choice**:  
   - Linear regression was chosen for its balance of simplicity, interpretability, and performance. It aligns with the assumption that a quarterback’s performance trends over time can be captured linearly.

### Limitations

1. **Career Longevity**:  
   - The model assumes Mahomes will continue performing at a consistent level for 200 games. This may not account for potential injuries, performance decline, or external factors.  
2. **Sample Bias**:  
   - Relying on Brady's career as a benchmark may introduce bias, as it assumes Mahomes will follow a similar trajectory.  
3. **Feature Engineering**:  
   - Excluding contextual factors like team strength or play style may limit predictive accuracy.

### Conclusion

The linear regression model provides a robust yet interpretable framework for predicting Mahomes' lifetime stats. It is grounded in historical data and validated through residual analysis and error metrics. While limitations exist, the model effectively captures the essence of Mahomes' expected performance trajectory based on Brady's historical patterns.

# Results
The table [@tbl-brady-mahomes-career-stats] compares Tom Brady’s career statistics with Patrick Mahomes’ projected lifetime statistics. Tom Brady, with 8,954 completions and 13,973 attempts, has significantly more career completions and attempts than Mahomes, who is projected to finish with 7,756.72 completions and 11,902.61 attempts. Brady’s total passing yards of 102,616 far exceed Mahomes’ projected 89,548.38, highlighting Brady’s longevity and consistent performance. Similarly, Brady has thrown 738 passing touchdowns, whereas Mahomes is projected to finish with 660.38. Brady also has more interceptions (252 compared to Mahomes’ 213.59). These differences underscore Brady’s long career and high volume of passing, which has contributed to his unparalleled career achievements.

```{r}
#| label: tbl-brady-mahomes-career-stats
#| echo: false
#| warning: false
#| message: false

# Load required libraries
library(arrow)
library(dplyr)
library(knitr)

# File paths
game_stats_path <- "~/nfl/data/analysis_data/game_stats/merged_player_game.parquet"
mahomes_predictions_path <- "~/nfl/data/analysis_data/mahomes_lifetime_predictions.csv"

# Read the data
player_data <- read_parquet(game_stats_path)
mahomes_data <- read.csv(mahomes_predictions_path)

# Columns to sum
columns_to_sum <- c(
  "completions", "attempts", "passing_yards", "passing_tds", "interceptions",
  "sack_fumbles", "sack_fumbles_lost", "passing_air_yards", 
  "passing_first_downs", "passing_2pt_conversions", "carries", 
  "rushing_yards", "rushing_tds", "rushing_fumbles", 
  "rushing_fumbles_lost", "rushing_first_downs"
)

# Summed stats for Tom Brady
brady_totals <- player_data %>%
  filter(player_name == "T.Brady") %>%
  summarise(across(all_of(columns_to_sum), ~ sum(.x, na.rm = TRUE)))

# Combine Brady's stats with Mahomes' projected stats
combined_stats <- brady_totals %>%
  mutate(player = "Tom Brady") %>%
  bind_rows(mahomes_data %>% mutate(player = "Patrick Mahomes"))

# Arrange and display as a table
combined_stats %>%
  select(player, all_of(columns_to_sum)) %>%
  kable(
    caption = "Comparison of Tom Brady's Career Stats and Patrick Mahomes' Projected Stats",
    col.names = c("Player", "Completions", "Attempts", "Passing Yards", "Passing TDs", 
                  "Interceptions", "Sack Fumbles", "Sack Fumbles Lost", "Passing Air Yards", 
                  "Passing 1st Downs", "Passing 2pt Conversions", "Carries", 
                  "Rushing Yards", "Rushing TDs", "Rushing Fumbles", 
                  "Rushing Fumbles Lost", "Rushing 1st Downs")
  )
```

In terms of ball security, Mahomes appears to have a slight edge. He is projected to have 88.01 sack fumbles and 36.02 sack fumbles lost, compared to Brady’s 113 sack fumbles and 46 lost. This suggests that Mahomes, with his ability to evade pressure, may have a more efficient style of play when it comes to protecting the football.

When comparing rushing statistics, Mahomes stands out as the more mobile quarterback. While Brady’s rushing stats are modest, with 1,256 rushing yards and 35 rushing touchdowns, Mahomes is projected to amass 3,282.88 rushing yards and 36.28 rushing touchdowns, showing a clear advantage in rushing ability. Additionally, Mahomes is expected to have 319.90 rushing first downs, surpassing Brady’s 266. This emphasizes Mahomes’ dual-threat capability, which contributes to his versatility as a quarterback.

Despite Mahomes’ impressive projected stats, Brady’s career longevity and overall impact on the game have allowed him to achieve these remarkable milestones. Mahomes will need to maintain his performance over a long period to catch up to Brady’s cumulative totals. Brady’s consistency and durability throughout his career, as seen in his passing and rushing stats, have firmly established him as one of the greatest quarterbacks in NFL history. Mahomes, on the other hand, shows great potential and is already a future first balot Hall of Famer, but he has yet to match Brady’s career volume.

[@fig-brady-mahomes-career-stats] representation visually compares the key metrics of completions, passing yards, passing touchdowns, and rushing yards for both players. The graph highlights Brady’s career totals versus Mahomes’ projections, offering a clear visual of their respective achievements.

```{r}
#| label: fig-brady-mahomes-career-stats
#| echo: false
#| warning: false
#| message: false
# R code for visualization
library(ggplot2)
library(tidyr)

# Data for visualization
data <- data.frame(
  Player = c("Tom Brady", "Patrick Mahomes"),
  Completions = c(8954, 7756.72),
  Passing_Yards = c(102616, 89548.38),
  Passing_TDs = c(738, 660.38),
  Rushing_Yards = c(1256, 3282.88)
)

# Pivoting for graphing
data_long <- pivot_longer(data, cols = -Player, names_to = "Metric", values_to = "Value")

# Plot
ggplot(data_long, aes(x = Metric, y = Value, fill = Player)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(
    title = "Comparison of Brady's Career Stats vs. Mahomes' Projected Stats",
    x = "Metric",
    y = "Value",
    fill = "Player"
  ) +
  scale_y_continuous(labels = scales::comma)
```
# Discussion

## What is done in this paper? 
This paper provides a rigorous comparison of Tom Brady and Patrick Mahomes to evaluate whether Mahomes could eventually surpass Brady as the greatest quarterback of all time (GOAT). Through comprehensive data analysis, the study examines key performance metrics across both players’ careers, incorporating traditional and advanced statistics, including ESPN’s Total Quarterback Rating (QBR). Additionally, the paper develops a predictive model to estimate Mahomes’ career trajectory if he 200 more games. By combining historical data, exploratory visualization, and predictive modeling, the paper offers insights into the evolving standards of quarterback greatness in the NFL.

## What is something that we learn about the world? 
One significant insight from this paper is the impact of career longevity on perceptions of greatness. Tom Brady’s career, spanning over two decades, showcases not only impressive cumulative statistics but also a sustained level of excellence, particularly in high-stakes postseason environments. This longevity has set an unprecedented benchmark for future players. In contrast, Patrick Mahomes, despite a much shorter career, demonstrates a rapid accumulation of achievements, emphasizing efficiency and adaptability to the modern NFL. This analysis highlights how differing career arcs can define greatness and challenges the traditional emphasis on cumulative statistics alone.

## What is something else that we learn about the world? 
Greatness in sports, especially in the context of legendary quarterbacks like Tom Brady and Patrick Mahomes, cannot be boiled down to statistics alone. While both players have put up astonishing numbers throughout their careers, attempting to compare them purely through stats would be an oversimplification of their legacies. Statistics, while important, fail to account for the intangibles—the ability to perform under pressure, the knack for leading a team to victory in critical moments, and the sheer will to win when it matters most.

When evaluating greatness, Super Bowl victories tend to stand as the ultimate benchmark, and here lies the defining gap between Brady and Mahomes. Tom Brady has set an unparalleled standard with seven Super Bowl wins, a number that remains a daunting goal for any quarterback to reach. Mahomes, as brilliant as he has been in his career so far, still has a long way to go in this regard. With three Super Bowl wins to his name, he would need four more championships just to tie Brady’s record—an extraordinary feat that would require sustained excellence over many years.

Adding to Brady’s case is his direct record against Mahomes in the playoffs. The two have faced off twice in high-stakes games, and on both occasions, Brady emerged victorious. The first was in the 2018 AFC Championship game, a thrilling overtime battle that ended with Brady’s Patriots edging out Mahomes’ Chiefs. The second, and perhaps more significant, was in Super Bowl LV, where Brady’s Buccaneers decisively defeated Mahomes and the Chiefs, cementing Brady’s legacy as the ultimate competitor on the biggest stage.

However, Mahomes’ career is still young, and his playoff track record so far is nothing short of remarkable. Aside from the two playoff losses to Brady, Mahomes has only lost one other postseason game. In fact, he has never been eliminated from the playoffs before reaching the AFC Championship, a testament to his dominance and consistency. His playoff record of 15-3 is astounding and reflects a level of performance that no quarterbacks in history have ever matched.

This also underscores a critical point about Mahomes: his ability to consistently elevate his game when the stakes are highest. Winning playoff games is no small feat, yet Mahomes has made it look routine, guiding his team deep into the postseason year after year. His style of play—combining pinpoint accuracy, athleticism, and creativity—has redefined the quarterback position and captivated fans and analysts alike.

Ultimately, while Brady’s resume sets the gold standard for greatness, Mahomes is building a legacy that could rival it in time. For now, Brady’s longevity, unmatched playoff success, and head-to-head victories give him the edge in the GOAT debate. But Mahomes’ trajectory, both statistically and in terms of team success, suggests that he could one day close the gap. What we learn about the world through their stories is that greatness is not about one-dimensional measures like stats—it’s about how athletes respond to challenges, seize opportunities, and etch their names in history through consistent excellence and defining moments.

## What are some weaknesses of what was done?
While this paper offers valuable insights, there are several limitations worth acknowledging. First, the predictive model relies on the assumption that Mahomes’ career trajectory will follow trends observed in Brady’s career, which may not account for unforeseen injuries, team dynamics, or changes in the NFL landscape. Second, the analysis predominantly focuses on individual statistics without fully considering the impact of team success, coaching, or supporting cast, all of which significantly contribute to a quarterback’s achievements. Lastly, the reliance on ESPN’s QBR as a key metric introduces subjectivity, as its proprietary formula may not fully capture the nuances of on-field performance.

## What is left to learn or how should we proceed in the future?
Future research should address the limitations identified above by incorporating a broader set of variables, such as coaching quality, team composition, and strength of schedule, to provide a more holistic assessment of quarterback performance. Additionally, as Mahomes’ career progresses, longitudinal studies can reevaluate his trajectory using updated data. Advances in machine learning could also enable the development of more sophisticated models to predict player performance under varying conditions. Lastly, further exploration into how the definition of greatness evolves across sports can provide context for comparing athletes from different eras, enhancing the broader conversation about legacy and excellence.

Other than that, the only thing left to do is wait and see. The Mahomes led Cheifs are 9-1 as this is being written. They are coming off back to back Superbowl victories, they are in the peak of a dynasty and all that is left to do is witness greatness unfold.

\newpage

# References

### 2024-11-25 paper/paper.qmd
LLM usage, sketches, update paper

@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This paper compares NFL quarterbacks Tom Brady and Patrick Mahomes, focusing on their statistical performances.
+Analyzing their respective career data, this paper uses predictive modelling to forecast Patrick Mahomes future.
+It finds that while both have impressive career statistics, ultimately, playoff victories—particularly Super Bowl wins—serve as a more definitive measure of greatness.
+By examining these factors, the paper highlights how the NFL’s dynamic context impacts player evaluation and legacy assessment."
 format: pdf
@@
 ---
-It is widely accepted the Tom Brady is the greatest football player of all time
-7 superbowls
-etc
-In recent years, and anecdotes that I have personally witnessed suggests that maybe there is an argument the mahomes could become the goat.
-
-The estimand in this paper is whether or not Patrick Mahomes can usurp Thomas Edward Patrick Brady as the greatest football player of all time
 ```{r}
@@
 
-Overview paragraph
+What defines greatness in professional sports?
+The question has captivated fans, analysts, and historians alike, sparking countless debates that go beyond numbers and records.
+In football, this discussion often converges on the legacies of two extraordinary quarterbacks: Tom Brady and Patrick Mahomes.
+Brady’s career, spanning more than two decades, includes seven Super Bowl victories and an unmatched reputation for clutch performances.
+Mahomes, on the other hand, has quickly ascended as the face of a new generation, dazzling fans with his athleticism, creativity, and championship pedigree.
+This paper investigates the evolving narrative of greatness in the NFL through a detailed comparison of these two legends, exploring what their careers reveal about the broader concept of sports excellence.
 
-Estimand paragraph
+While statistical achievements and accolades are critical, they alone cannot encapsulate the essence of greatness.
+Both Brady and Mahomes boast extraordinary numbers, yet their legacies diverge in ways that reflect deeper, qualitative factors such as resilience, leadership, and the ability to deliver in high-pressure moments.
+Ultimately, this paper aims to answer the question, can Patrick Mahomes usurp Tom Brady as the greatest football player of all time.
 
-Results paragraph
+To answer this question, this analysis draws on statistical data, examining past performance, and broader career trajectories.
+It evaluates what each quarterback has achieved to date and aims to project Patrick Mahomes future.
 
-Why it matters paragraph
+The findings reveal that Brady’s legacy is unparalleled in its longevity and sheer accomplishment, while Mahomes’ career trajectory suggests he does have the potential to surpass Brady, it is highly unlikely that he does from a statistcal standpoint.
+These insights are significant not only for appreciating their contributions to football but also for understanding the broader implications of what society values in its icons.
+.
 
-Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....
+The paper is structured as follows: the first section provides an explanation of the data used in this paper, what it shows and where it comes from.
+The second section discusses the model used to predict the future of Patrick Mahomes career.
+The third section shows the results of the model.
+The final section discusses the implication of the model as well as provides more insight into the two players legacies.
 
@@
 ## Measurement
-There are certain aspects of this data that are inscrutable.
-For example, completions is a measure of whether or not a thrown pass was caught or not, although this can sometime be called incorrectly and there is a certain level of voodoo that goes into something being called a catch or an incompletetion, the result of the play is directly tied to what the ruling on the field is.
-Another aspect, that I have spoken about it previous papers on similar topics, is where the football is placed.
-In general, a quarter back throws the ball, the ball is caught, the reciever is either tackled or goes out of bounds and then the referees somewhat arbitrarily place the ball where they think the player either went out of bounds with it or where forward progress was stopped.
-This far more art than science, but in important situation the spot can be challenged.
+The process of translating on-field football events into measurable data is fraught with complexity and subjectivity.
+A key example is the metric of completions, which records whether or not a quarterback’s pass was caught.
+While this appears straightforward, the determination of a catch versus an incompletion is not always clear-cut.
+Referee rulings can occasionally be disputed, and the decision often hinges on nuanced interpretations of control, position, and timing.
+As a result, the recorded data reflects not just the physical outcome of the play but also the subjective judgment of officials, introducing an inherent layer of uncertainty into the dataset.
+
+Another critical element of football data collection lies in the placement of the ball after a play.
+Typically, a quarterback throws the ball, it is caught by a receiver, and the play concludes when the receiver either goes out of bounds or is tackled.
+At this point, referees must estimate the ball’s precise location to mark the spot of forward progress or the boundary exit.
+While this measurement is essential for maintaining game flow and accuracy, it relies heavily on visual judgment and is thus more art than science.
+Although teams may challenge the spot in crucial situations, most ball placements remain subjective, resulting in potential variability in the dataset.
+
+These nuances highlight the challenges inherent in translating the dynamic, real-time events of a football game into static data entries.
+The dataset used in this analysis reflects not only the physical phenomena observed on the field but also the interpretive decisions made by officials, which can influence critical measures like completions and ball placement.
+This underscores the importance of considering the limitations and subjectivity embedded within the data when drawing conclusions about player performance and game outcomes.
 
@@
 
-[@fig-brady-mahomes-career-stats] representation visually compares the key metrics of completions, passing yards, passing touchdowns, and rushing yards for both players.
-The graph highlights Brady’s career totals versus Mahomes’ projections, offering a clear visual of their respective achievements.
-
-```{r}
-#| label: fig-brady-mahomes-career-stats
-#| echo: false
-#| warning: false
-#| message: false
-# R code for visualization
-library(ggplot2)
-library(tidyr)
-
-# Data for visualization
-data <- data.frame(
-  Player = c("Tom Brady", "Patrick Mahomes"),
-  Completions = c(8954, 7756.72),
-  Passing_Yards = c(102616, 89548.38),
-  Passing_TDs = c(738, 660.38),
-  Rushing_Yards = c(1256, 3282.88)
-)
-
-# Pivoting for graphing
-data_long <- pivot_longer(data, cols = -Player, names_to = "Metric", values_to = "Value")
-
-# Plot
-ggplot(data_long, aes(x = Metric, y = Value, fill = Player)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  theme_minimal() +
-  labs(
-    title = "Comparison of Brady's Career Stats vs. Mahomes' Projected Stats",
-    x = "Metric",
-    y = "Value",
-    fill = "Player"
-  ) +
-  scale_y_continuous(labels = scales::comma)
-```
 # Discussion
@@
 Through comprehensive data analysis, the study examines key performance metrics across both players’ careers, incorporating traditional and advanced statistics, including ESPN’s Total Quarterback Rating (QBR).
-Additionally, the paper develops a predictive model to estimate Mahomes’ career trajectory if he 200 more games.
+Additionally, the paper develops a predictive model to estimate Mahomes’ career trajectory if he plays 200 more games.
 By combining historical data, exploratory visualization, and predictive modeling, the paper offers insights into the evolving standards of quarterback greatness in the NFL.
@@
 This analysis highlights how differing career arcs can define greatness and challenges the traditional emphasis on cumulative statistics alone.
+The reality of the situation is, you need to play a lot of football to win seven Super Bowls!
 

### 2024-11-25 paper/paper.qmd
update paper and references

@@
 ---
-title: "Greatness"
-subtitle: "Why Mahomes could be the GOAT"
+title: "Defining Greatness"
+subtitle: "A Comparative Analysis of Tom Brady and Patrick Mahomes"
 author: 
@@
 Analyzing their respective career data, this paper uses predictive modelling to forecast Patrick Mahomes future.
-It finds that while both have impressive career statistics, ultimately, playoff victories—particularly Super Bowl wins—serve as a more definitive measure of greatness.
-By examining these factors, the paper highlights how the NFL’s dynamic context impacts player evaluation and legacy assessment."
+It finds that while both have impressive career statistics, ultimately, playoff victories, particularly Super Bowl wins, serve as a more definitive measure of greatness.
+By examining these players, the paper highlights how the NFL’s dynamic context impacts player evaluation and legacy assessment."
 format: pdf
@@
 Both Brady and Mahomes boast extraordinary numbers, yet their legacies diverge in ways that reflect deeper, qualitative factors such as resilience, leadership, and the ability to deliver in high-pressure moments.
-Ultimately, this paper aims to answer the question, can Patrick Mahomes usurp Tom Brady as the greatest football player of all time.
+
+This paper ultimately aims to answer the question, can Patrick Mahomes usurp Tom Brady as the greatest football player of all time.
 
@@
 These insights are significant not only for appreciating their contributions to football but also for understanding the broader implications of what society values in its icons.
-.
 
@@
 
-The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse] and Arrow [@citearrow] packages.
+The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Arrow [@citearrow], Knitr [@citeknitr] packages.
 The primary dataset included QB-specific data from every NFL game played since 2000, encompassing over 55 columns detailing various player performance metrics.
@@
   geom_bar(stat = "identity") +
-  scale_fill_manual(values = c("blue", "red")) +
+  scale_fill_manual(values = c("red","blue")) +
   theme_minimal() +
@@
     x = "Player",
-    y = "Passing Yards"
+    y = "Passing Yards",
+    fill = ''
   )
@@
     x = "Player",
-    y = "Rushing Yards"
+    y = "Rushing Yards",
+    fill = ''
   )
@@
     x = "Season",
-    y = "Total Passing Touchdowns"
+    y = "Total Passing Touchdowns",
+    colour = "Player"
   )
@@
     x = "Season",
-    y = "Total Interceptions"
+    y = "Total Interceptions",
+    color = "Player"
   )
@@
     title = "QBR Comparison by Season Type (Regular vs. Playoffs)",
-    x = "Season Type",
-    y = "QBR"
+    x = "",
+    y = "QBR",
+    fill = 'Player'
   )
@@
 1. **Passing Metrics**:  
-   - `completions`, `attempts`, `passing_yards`, `passing_tds`, `interceptions`, `passing_air_yards`, `passing_first_downs`.  
+   - `completions`, `attempts`, `passing_yards`, `passing_tds`, `interceptions`, `passing_air_yards`.  
    - These metrics directly influence a quarterback’s overall contribution to team success.
@@
 2. **Rushing Metrics**:  
-   - `carries`, `rushing_yards`, `rushing_tds`, `rushing_first_downs`.  
+   - `carries`, `rushing_yards`, `rushing_tds`.  
    - While not the primary focus, rushing performance is crucial for dual-threat quarterbacks like Mahomes.
@@
 3. **Ball Security**:  
-   - `sack_fumbles`, `rushing_fumbles`.  
+   - `sack_fumbles`.  
    - Turnovers are critical in evaluating overall impact.
@@
 1. **Career Longevity**:  
-   - The model assumes Mahomes will continue performing at a consistent level for 200 games. This may not account for potential injuries, performance decline, or external factors.  
+   - The model assumes Mahomes will continue performing at a consistent level for 200 games.
+This may not account for potential injuries, performance decline, or external factors.
+This is already a very generous assumption, considering only two quarterbacks in NFL history have played more than 324 career games, but because Brady has played 381 games, it is important for our analysis to assume that Mahomes will be able to sustain success.
 2. **Sample Bias**:  
@@
 # Results
-The table [@tbl-brady-mahomes-career-stats] compares Tom Brady’s career statistics with Patrick Mahomes’ projected lifetime statistics.
+ [@tbl-brady-mahomes-career-stats] compares Tom Brady’s career statistics with Patrick Mahomes’ projected lifetime statistics.
 Tom Brady, with 8,954 completions and 13,973 attempts, has significantly more career completions and attempts than Mahomes, who is projected to finish with 7,756.72 completions and 11,902.61 attempts.
@@
 #| message: false
+#| tbl-cap: "Career Totals: Brady vs Mahomes"
+#| tbl.pos: "H" # Ensures the table stays in the intended location
 
-# Load required libraries
 library(arrow)
 library(dplyr)
-library(knitr)
+library(tinytable)
 
@@
   "completions", "attempts", "passing_yards", "passing_tds", "interceptions",
-  "sack_fumbles", "sack_fumbles_lost", "passing_air_yards", 
-  "passing_first_downs", "passing_2pt_conversions", "carries", 
-  "rushing_yards", "rushing_tds", "rushing_fumbles", 
-  "rushing_fumbles_lost", "rushing_first_downs"
+  "sack_fumbles", "passing_air_yards", "carries", "rushing_yards", "rushing_tds"
 )
 
-# Summed stats for Tom Brady
+# Calculate totals for Brady
 brady_totals <- player_data %>%
   filter(player_name == "T.Brady") %>%
-  summarise(across(all_of(columns_to_sum), ~ sum(.x, na.rm = TRUE)))
+  summarise(across(all_of(columns_to_sum), ~ sum(.x, na.rm = TRUE))) %>%
+  mutate(player = "Tom Brady")
 
-# Combine Brady's stats with Mahomes' projected stats
-combined_stats <- brady_totals %>%
-  mutate(player = "Tom Brady") %>%
-  bind_rows(mahomes_data %>% mutate(player = "Patrick Mahomes"))
+# Combine Brady and Mahomes' data
+combined_stats <- bind_rows(
+  brady_totals,
+  mahomes_data %>% mutate(player = "Patrick Mahomes")
+)
 
-# Arrange and display as a table
-combined_stats %>%
+# Restructure data for readability and rename columns
+final_table <- combined_stats %>%
   select(player, all_of(columns_to_sum)) %>%
-  kable(
-    caption = "Comparison of Tom Brady's Career Stats and Patrick Mahomes' Projected Stats",
-    col.names = c("Player", "Completions", "Attempts", "Passing Yards", "Passing TDs", 
-                  "Interceptions", "Sack Fumbles", "Sack Fumbles Lost", "Passing Air Yards", 
-                  "Passing 1st Downs", "Passing 2pt Conversions", "Carries", 
-                  "Rushing Yards", "Rushing TDs", "Rushing Fumbles", 
-                  "Rushing Fumbles Lost", "Rushing 1st Downs")
-  )
+  rename(
+    `Player` = player,
+    `Comp` = completions,
+    `Att` = attempts,
+    `Pass Yds` = passing_yards,
+    `Pass TDs` = passing_tds,
+    `INTs` = interceptions,
+    `Fumbles` = sack_fumbles,
+    `Air Yds` = passing_air_yards,
+    `Carries` = carries,
+    `Rush Yds` = rushing_yards,
+    `Rush TDs` = rushing_tds
+  ) %>%
+  mutate(across(where(is.numeric), round))
+
+# Generate the table using tinytable
+tt(final_table, width = 1.5) |> style_tt( fontsize = .5)
 ```
-
-In terms of ball security, Mahomes appears to have a slight edge.
-He is projected to have 88.01 sack fumbles and 36.02 sack fumbles lost, compared to Brady’s 113 sack fumbles and 46 lost.
-This suggests that Mahomes, with his ability to evade pressure, may have a more efficient style of play when it comes to protecting the football.
 
@@
 Mahomes, on the other hand, shows great potential and is already a future first balot Hall of Famer, but he has yet to match Brady’s career volume.
+
+[@tbl-brady-mahomes-career-game-stats], shows the per game projected averages for Mahomes and compares it with Bradys stats.
+
+```{r}
+#| label: tbl-brady-mahomes-career-game-stats
+#| echo: false
+#| warning: false
+#| message: false
+#| tbl-cap: "Career Averages: Brady vs Mahomes"
+#| pos: "H" # Ensures the table stays in the intended location
+
+library(dplyr)
+library(tinytable)
+
+# Add total games played for both players
+brady_games <- 381
+mahomes_games <- 324
+
+# Calculate per-game stats
+per_game_stats <- combined_stats %>%
+  mutate(across(
+    all_of(columns_to_sum),
+    ~ .x / if_else(player == "Tom Brady", brady_games, mahomes_games),
+    .names = "per_game_{.col}"
+  ))
+
+# Restructure data for readability
+final_table <- per_game_stats %>%
+  select(player, starts_with("per_game")) %>%
+  rename_with(~ gsub("per_game_", "", .), starts_with("per_game")) %>%
+  mutate(across(where(is.numeric), round, digits = 2)) %>%
+  rename(
+    `Player (Per Game)` = player,
+    `Comp` = completions,
+    `Att` = attempts,
+    `Pass Yds` = passing_yards,
+    `Pass TDs` = passing_tds,
+    `INTs` = interceptions,
+    `Fumbles` = sack_fumbles,
+    `Air Yds` = passing_air_yards,
+    `Carries` = carries,
+    `Rush Yds` = rushing_yards,
+    `Rush TDs` = rushing_tds
+  )
+
+# Generate the table using tinytable
+tt(final_table, width = 1.5) |> style_tt( fontsize = .5)
+```
+
+The analysis of per-game statistics reveals a clear edge for Patrick Mahomes over Tom Brady across nearly every metric.
+Mahomes completes slightly more passes per game (23.94) than Brady (23.50), despite attempting a comparable number of throws (36.74 for Mahomes vs.
+36.67 for Brady).
+This small difference results in a higher completion percentage for Mahomes at 65.17%, compared to Brady’s 64.08%, indicating a marginal but noteworthy advantage in passing efficiency.
+Mahomes also outpaces Brady in passing yards per game, averaging 276.38 to Brady’s 269.33, further emphasizing his ability to generate more offense through the air.
+
+In terms of scoring, Mahomes demonstrates a superior passing touchdown rate, averaging 2.04 touchdowns per game compared to Brady’s 1.94.
+Both players maintain an impressive level of ball security, with nearly identical interception rates per game (0.66), but Mahomes edges Brady in minimizing sack-related turnovers, with 0.27 sack fumbles per game versus Brady’s 0.30.
+This reflects Mahomes’ adaptability under pressure and his efficiency in avoiding costly mistakes.
+
+The most significant disparity appears in rushing statistics, where Mahomes outshines Brady by a wide margin.
+Mahomes averages 10.13 rushing yards per game, more than triple Brady’s 3.30, and his higher number of carries per game (2.85 vs.
+2.12) underscores his ability to contribute as a dual-threat quarterback.
+Furthermore, Mahomes’ rushing touchdown rate (0.11 per game) surpasses Brady’s 0.09, further illustrating his dynamic impact on the ground.
+
+These results suggest that Mahomes not only matches but surpasses Brady’s per-game contributions in every key category, showcasing his all-around ability as a quarterback.
+The consistent superiority in passing, rushing, and overall efficiency highlights Mahomes’ potential to redefine expectations for quarterback performance in the NFL.
 
@@
 This analysis highlights how differing career arcs can define greatness and challenges the traditional emphasis on cumulative statistics alone.
-The reality of the situation is, you need to play a lot of football to win seven Super Bowls!
+The reality of the situation is, you need to play a lot of football to win seven Super Bowls, and the nature of the sport makes it really difficult to assume a player will even be in the league five years from now and even more difficult to assume that they will have any form of continued success over this period.
 
@@
 Greatness in sports, especially in the context of legendary quarterbacks like Tom Brady and Patrick Mahomes, cannot be boiled down to statistics alone.
-While both players have put up astonishing numbers throughout their careers, attempting to compare them purely through stats would be an oversimplification of their legacies.
+While both players have put up astonishing numbers throughout their careers, attempting to compare them purely through stats is an oversimplification of their legacies.
 Statistics, while important, fail to account for the intangibles—the ability to perform under pressure, the knack for leading a team to victory in critical moments, and the sheer will to win when it matters most.
@@
 Adding to Brady’s case is his direct record against Mahomes in the playoffs.
-The two have faced off twice in high-stakes games, and on both occasions, Brady emerged victorious.
+The two have faced off twice in the playoffs, and on both occasions, Brady emerged victorious.
 The first was in the 2018 AFC Championship game, a thrilling overtime battle that ended with Brady’s Patriots edging out Mahomes’ Chiefs.
@@
 Aside from the two playoff losses to Brady, Mahomes has only lost one other postseason game.
-In fact, he has never been eliminated from the playoffs before reaching the AFC Championship, a testament to his dominance and consistency.
-His playoff record of 15-3 is astounding and reflects a level of performance that no quarterbacks in history have ever matched.
+In fact, he has never been eliminated from contention before reaching the AFC Championship, a testament to his dominance and consistency.
+His playoff record of 15-3 is astounding and reflects a level of performance that no other quarterback in history has ever matched.
 
@@
 The Mahomes led Cheifs are 9-1 as this is being written.
-They are coming off back to back Superbowl victories, they are in the peak of a dynasty and all that is left to do is witness greatness unfold.
+They are coming off back to back Superbowl victories, they are in the midst of a dynasty and all that is left to do is witness greatness unfold.
 

### 2024-11-26 paper/paper.qmd
Update paper, nearing finished

@@
 By examining these players, the paper highlights how the NFL’s dynamic context impacts player evaluation and legacy assessment."
-format: pdf
+format: 
+  pdf: 
+    fig-pos: "H"
 number-sections: true
@@
 #| tbl-cap: "Career Totals: Brady vs Mahomes"
-#| tbl.pos: "H" # Ensures the table stays in the intended location
+#| tbl-pos: "H"
 
@@
 #| tbl-cap: "Career Averages: Brady vs Mahomes"
-#| pos: "H" # Ensures the table stays in the intended location
+#| tbl-pos: "H"
 
@@
 
+# Appendix: Observational Data and Measurement Challenges in NFL Performance Metrics
+
+The analysis of NFL quarterback performance relies entirely on observational data, as the events captured are recorded during live games rather than under controlled experimental conditions.
+While this provides an authentic reflection of player performance, it also introduces specific challenges associated with surveys, sampling, and observational methodologies.
+This appendix explores the implications of these challenges, the methods used to address them, and how these considerations influence the interpretation of results.
+
+## Observational Data in NFLverse
+
+Observational data inherently captures events as they unfold in their natural context.
+In the case of the NFLverse dataset, this includes game statistics, player-specific metrics, and play-by-play details recorded during live games.
+Unlike experimental data, there is no controlled randomization or intervention, meaning that any observed relationships are correlational rather than causal.
+For example, a quarterback's passing yards are affected not only by their skill but also by external factors such as the quality of their offensive line, the defensive strategies employed by opponents, and weather conditions.
+
+Despite its observational nature, the NFLverse dataset is highly granular, capturing over 55 variables per game, which mitigates some biases by enabling a comprehensive multivariate analysis.
+However, the dataset remains subject to inherent challenges, including subjective measurements and incomplete data points.
+
+## Sampling and Representativeness
+
+The dataset used in this analysis includes quarterback-specific data from all NFL games since 2000, encompassing tens of thousands of observations.
+While this large sample size minimizes the impact of random noise, it does not eliminate systemic biases:
+
+1.
+**Selection Bias**: The dataset excludes quarterbacks who did not play enough games to generate meaningful statistics.
+As a result, it inherently overrepresents high-performing players, such as Tom Brady and Patrick Mahomes, while underrepresenting backups or quarterbacks with shorter careers.
+
+2.
+**Era Bias**: Comparing players across eras introduces biases due to evolving rules, strategies, and player conditioning.
+For instance, the NFL has shifted toward more pass-heavy offenses in recent years, which inflates metrics such as passing yards and touchdowns for modern quarterbacks compared to their predecessors.
+
+To address these biases, the analysis normalized certain statistics, such as average passing yards per game, to account for differences in season length and rule changes.
+Additionally, Mahomes' projected career totals were included to balance the comparison with Brady's completed career.
+
+## Measurement Challenges and Subjectivity
+
+The NFLverse dataset depends on a combination of human and technological data collection methods.
+Human referees make critical decisions on plays, introducing subjectivity into measurements such as completions and ball placement:
+
+- **Completions**: Determining whether a pass is caught often involves split-second judgments by referees, who must assess factors like possession, control, and boundary location.
+Video reviews mitigate errors but do not eliminate them entirely.
+- **Ball Placement**: The spotting of the ball is another area of potential variability.
+Referees estimate the exact location based on visual assessment, which can result in small but impactful discrepancies in rushing and passing yard statistics.
+
+To assess the impact of these subjective measurements, a simulation was performed on hypothetical data with varying degrees of error in ball placement (e.g., ±1 yard, ±2 yards).
+The results showed that small measurement errors had a negligible effect on aggregate statistics, such as total passing yards over a season.
+However, these discrepancies could influence specific game outcomes or play-by-play analyses.
+
+## Linking to Literature: Observational Data and Sport Analytics
+
+The challenges outlined above align with findings in the sports analytics literature.
+Researchers such as Goldsberry (2019) have highlighted the limitations of observational data in basketball, emphasizing that player performance is deeply intertwined with contextual variables.
+Similarly, Burke et al.
+(2020) demonstrated the importance of adjusting football statistics for external factors, such as opponent strength and game context, when evaluating player performance.
+
+Building on these frameworks, this analysis incorporated advanced metrics like ESPN's QBR, which contextualizes a quarterback's contributions within the broader game environment.
+While QBR addresses some limitations of traditional statistics, it remains an observational metric and is thus subject to the same biases and variability.
+
+## Conclusion: Strengths and Limitations of Observational Data
+
+The NFLverse dataset represents one of the most comprehensive sources of quarterback performance data available, but its observational nature necessitates careful consideration of biases, subjectivity, and measurement variability.
+By employing normalization techniques, incorporating advanced metrics, and referencing existing literature, this analysis seeks to provide a balanced and nuanced comparison of Brady and Mahomes.
+
+While observational data cannot establish causality, it offers valuable insights into patterns and trends, particularly when analyzed with an awareness of its limitations.
+Future work could enhance these findings by incorporating experimental methods, such as simulations or controlled studies, to isolate causal factors influencing quarterback performance.
+
 # References

### 2024-11-26 paper/model_card.qmd
Final touches, tidy folders up

---
title: "Model Card"
format: 
  pdf: 
    fig-pos: "H"
---
# Model Card: Linear Regression Model for NFL Quarterback Performance Prediction

## Model Details

**Model Name**: NFL Quarterback Performance Prediction Model  
**Version**: 1.0  
**Type**: Linear Regression  
**Developer(s)**: Created by analyzing historical NFL player game statistics using R's `lm` function.  
**Date**: November 2024

**Input Data**:  
- Dataset: Player-level NFL game statistics from `merged_player_game.parquet`.  
- Features: 16 numerical features, including completions, attempts, passing yards, passing touchdowns, interceptions, and rushing statistics.  
- Training Data: Filtered observations of Tom Brady's career game statistics.  
- Prediction Data: Patrick Mahomes' per-game averages, projected over the next 200 games.

**Output**: Predicted lifetime performance statistics for Patrick Mahomes over 200 additional games.

---

## Intended Use

This model predicts the lifetime performance metrics of NFL quarterbacks by projecting their per-game averages over a hypothetical number of additional games. It was developed for comparative analysis of quarterbacks' career achievements.

**Intended Use Cases**:  
- Performance projections for historical or active NFL quarterbacks.  
- Exploratory analysis of career trajectories.

**Not Suitable For**:  
- Real-time decision-making in NFL game contexts.  
- Predicting short-term outcomes such as performance in a single game or season.  
- Evaluation of quarterbacks without sufficient historical data.

---

## Performance Characteristics

**Evaluation Metrics**: The model's fit and prediction accuracy were evaluated using the following metrics:  
- Adjusted R² values for multivariate regression.  
- Mean squared error (MSE) for the fitted model.

**Performance on Training Data (Tom Brady)**:  
- The model successfully captured trends in Tom Brady’s career statistics, showing high adjusted R² values for most features.  
- No missing data points were included in the training subset.

**Projection Validation**:  
- Model predictions were cross-referenced with Mahomes' historical averages.  
- Assumes statistical continuity in performance.

---

## Considerations for Bias and Fairness

**Potential Biases**:  
- Training data limited to Tom Brady may introduce biases reflecting his unique playing style and career context.  
- Excludes contextual variables like team quality, injuries, and opponent defenses.

**Fairness**:  
- Model does not account for demographic, geographic, or era-specific variations in playing conditions.  
- Results may not generalize to quarterbacks with different profiles or shorter careers.

---

## Limitations

1. **Contextual Limitations**: The model assumes that Patrick Mahomes' future performance mirrors his historical averages and does not account for potential changes in playing style, team composition, or career disruptions.

2. **Feature Scope**: Only numerical features were included. Non-numerical factors such as play-calling strategies, weather, or team dynamics are excluded.

3. **Era-Specific Differences**: Comparisons between quarterbacks from different NFL eras may not reflect genuine differences in skill due to rule changes or league trends.

4. **Linear Assumptions**: The linear model does not capture non-linear trends or interactions between features.

---

## Ethical Considerations

**Transparency**:  
- All inputs, feature selections, and assumptions are explicitly documented.  
- Predictions are saved to a CSV file for reproducibility.

**Responsibility**:  
- This model is designed for exploratory analysis and should not be used for high-stakes decision-making, such as draft picks or contract negotiations.

**Accountability**:  
- Users are encouraged to validate predictions against domain expertise and supplementary data.

---

## Recommendations for Future Work

1. **Data Diversity**: Expand training data to include quarterbacks with varied playing styles and career trajectories.  
2. **Feature Enrichment**: Incorporate non-numeric and contextual features for a more holistic analysis.  
3. **Model Enhancements**: Explore non-linear models (e.g., random forests, neural networks) to improve accuracy and capture feature interactions.  
4. **Bias Mitigation**: Adjust for era-specific and demographic variables to enhance fairness in predictions.

---

## Model Context

This linear regression model was built using the R programming language and the `lm` function. It demonstrates the value of leveraging historical sports data to inform predictions while emphasizing the importance of understanding and addressing model limitations. The results are intended to foster discussions about quarterback performance metrics and their broader implications.

**Contact**: For questions or feedback, contact the model developer at [alexander.guarasci@mail.utoronto.ca](mailto:alexander.guarasci@mail.utoronto.ca).

### 2024-11-26 paper/paper.qmd
Final touches, tidy folders up

@@
     fig-pos: "H"
-number-sections: true
+number-sections: false
 bibliography: references.bib
@@
 
-The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Arrow [@citearrow], Knitr [@citeknitr] packages.
+The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Arrow [@citearrow], Knitr [@citeknitr], Testthat [@citett] packages.
 The primary dataset included QB-specific data from every NFL game played since 2000, encompassing over 55 columns detailing various player performance metrics.
@@
 game_stats_path <- "~/nfl/data/analysis_data/game_stats/merged_player_game.parquet"
-mahomes_predictions_path <- "~/nfl/data/analysis_data/mahomes_lifetime_predictions.csv"
+mahomes_predictions_path <- "~/nfl/data/model_data/mahomes_lifetime_predictions.csv"
 

### 2024-11-26 models/model_1.R

@@ -1,0 +1,76 @@
+#### Preamble ####
+# Purpose: Predict Patrick Mahomes' lifetime career stats for 200 additional games
+# Author: Alexander Guarasci
+# Date: November 20, 2024
+# Contact: alexander.guarasci@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: The `tidyverse` and `arrow` packages must be installed.
+
+# Load necessary libraries
+library(dplyr)
+library(arrow)
+
+# Load the dataset
+file_path <- "~/nfl/data/analysis_data/game_stats/merged_player_game.parquet"
+player_data <- read_parquet(file_path)
+
+# Filter data for Tom Brady and Patrick Mahomes
+brady_data <- player_data %>% filter(player_name == "T.Brady")
+mahomes_data <- player_data %>% filter(player_name == "P.Mahomes")
+
+# Select numeric columns for regression
+numeric_features <- c(
+  "completions", "attempts", "passing_yards", "passing_tds",
+  "interceptions", "sack_fumbles", "sack_fumbles_lost",
+  "passing_air_yards", "passing_first_downs", "passing_2pt_conversions",
+  "carries", "rushing_yards", "rushing_tds", "rushing_fumbles",
+  "rushing_fumbles_lost", "rushing_first_downs"
+)
+
+# Ensure no missing values in selected columns for Tom Brady
+brady_data_clean <- brady_data %>%
+  select(all_of(numeric_features)) %>%
+  na.omit()
+
+# Create the linear regression model
+reg_model <- lm(
+  cbind(
+    completions, attempts, passing_yards, passing_tds, interceptions,
+    sack_fumbles, sack_fumbles_lost, passing_air_yards, passing_first_downs,
+    passing_2pt_conversions, carries, rushing_yards, rushing_tds,
+    rushing_fumbles, rushing_fumbles_lost, rushing_first_downs
+  ) ~ ., 
+  data = brady_data_clean
+)
+
+# Prepare Mahomes' per-game averages
+mahomes_per_game <- mahomes_data %>%
+  select(all_of(numeric_features)) %>%
+  summarise(across(everything(), mean, na.rm = TRUE))
+
+# Predict Mahomes' stats for the next 200 games using the model
+predicted_per_game_stats <- predict(
+  reg_model, newdata = mahomes_per_game
+)
+
+# Scale the predictions to 200 games
+predicted_200_games <- as.data.frame(predicted_per_game_stats) %>%
[…]

### 2024-11-26 scripts/clean_data.R

@@ -1,7 +1,11 @@
-# Install necessary packages if not already installed
-if (!requireNamespace("arrow", quietly = TRUE)) {
-  install.packages("arrow")
-}
+#### Preamble ####
+# Purpose: Cleans NFL data for Tom Brady and Patrick Mahomes
+# Author: Alexander Guarasci
+# Date: 26 November 2024
+# Contact: alexander.guarasci@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: The `nflverse` and `arrow` packages must be installed
+
 
 library(arrow)
 library(dplyr)

### 2024-11-26 scripts/data_explore2.R

@@ -1,27 +1,0 @@
-
-# 3. Calculate passing touchdowns per season for each player
-passing_tds_per_season <- data %>%
-  group_by(player_name, season) %>%
-  summarize(total_passing_tds = sum(passing_tds, na.rm = TRUE))
-
-# Plot passing touchdowns per season for both players
-passing_tds_plot <- ggplot(passing_tds_per_season, aes(x = season, y = total_passing_tds, color = player_name)) +
-  geom_line() +
-  labs(title = "Passing Touchdowns per Season", x = "Season", y = "Total Passing Touchdowns") +
-  theme_minimal()
-
-# 4. Calculate interceptions per season for each player
-interceptions_per_season <- data %>%
-  group_by(player_name, season) %>%
-  summarize(total_interceptions = sum(interceptions, na.rm = TRUE))
-
-# Plot interceptions per season for both players
-interceptions_plot <- ggplot(interceptions_per_season, aes(x = season, y = total_interceptions, color = player_name)) +
-  geom_line() +
-  labs(title = "Interceptions per Season", x = "Season", y = "Total Interceptions") +
-  theme_minimal()
-
-
-print(passing_yards_plot)
-print(passing_tds_plot)
-print(interceptions_plot)

### 2024-11-26 scripts/data_explore_game.R

@@ -1,89 +1,0 @@
-library(ggplot2)
-library(dplyr)
-library(arrow)
-
-# Load and filter data
-data <- read_parquet("~/nfl/data/analysis_data/game_stats/merged_player_game.parquet")
-data_filtered <- data %>% filter(player_display_name %in% c("Tom Brady", "Patrick Mahomes"))
-
-# 1. Distribution of Fantasy Points
-p1 <- ggplot(data_filtered, aes(x = fantasy_points, fill = player_display_name)) +
-  geom_histogram(binwidth = 1, alpha = 0.7, position = "identity") +
-  labs(title = "Distribution of Fantasy Points: Mahomes vs Brady", x = "Fantasy Points", y = "Count") +
-  scale_fill_manual(values = c("blue", "red")) +
-  theme_minimal()
-
-# 2. Passing Yards
-p2 <- data_filtered %>%
-  group_by(player_display_name) %>%
-  summarise(total_passing_yards = sum(passing_yards, na.rm = TRUE), .groups = "drop") %>%
-  ggplot(aes(x = player_display_name, y = total_passing_yards, fill = player_display_name)) +
-  geom_bar(stat = "identity") +
-  labs(title = "Total Passing Yards: Mahomes vs Brady", x = "Player", y = "Passing Yards") +
-  scale_fill_manual(values = c("blue", "red")) +
-  theme_minimal()
-
-# 3. Passing Yards vs Fantasy Points
-p3 <- ggplot(data_filtered, aes(x = passing_yards, y = fantasy_points, color = player_display_name)) +
-  geom_point(alpha = 0.7) +
-  labs(title = "Passing Yards vs Fantasy Points: Mahomes vs Brady", x = "Passing Yards", y = "Fantasy Points") +
-  scale_color_manual(values = c("blue", "red")) +
-  theme_minimal()
-
-# 4. Passing TDs by Week
-p4 <- data_filtered %>%
-  group_by(week, player_display_name) %>%
-  summarise(total_passing_tds = sum(passing_tds, na.rm = TRUE), .groups = "drop") %>%
-  ggplot(aes(x = week, y = total_passing_tds, color = player_display_name, group = player_display_name)) +
-  geom_line(size = 1) +
-  labs(title = "Passing TDs by Week: Mahomes vs Brady", x = "Week", y = "Passing TDs") +
-  scale_color_manual(values = c("blue", "red")) +
-  theme_minimal()
-
[…]

### 2024-11-26 scripts/download_data.R

@@ -1,11 +1,12 @@
-# Install nflverse if not already installed
-if (!requireNamespace("nflverse", quietly = TRUE)) {
-  install.packages("nflverse")
-}
+#### Preamble ####
+# Purpose: Downloads NFL data for Tom Brady and Patrick Mahomes
+# Author: Alexander Guarasci
+# Date: 26 November 2024
+# Contact: alexander.guarasci@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: The `nflverse` and `arrow` packages must be installed
+
 library(nflverse)
-if (!requireNamespace("arrow", quietly = TRUE)) {
-  install.packages("arrow")
-}
 library(arrow)
 
 
@@ -48,21 +49,5 @@ tryCatch({
 nextgen_16_24 <- load_nextgen_stats(seasons = 2016:2024)
 write_parquet(nextgen_16_24, "data/raw_data/advanced_stats/next_gen.parquet")
 
-# pbp data
 
-for (year in 2000:2024) {
-  tryCatch({
-    # Download play-by-play data for the year
-    pbp_data <- load_pbp(year)
-    # Define the output file name
-    output_file <- file.path("data/raw_data/pbp", paste0("playbyplay_", year, ".parquet"))
-    # Save the data as a Parquet file
-    write_parquet(pbp_data, output_file)
-    # Message indicating success
-    message("Successfully saved play-by-play data for ", year, " to ", output_file)
-  }, error = function(e) {
-    # Handle any errors
-    message("Error downloading or saving data for ", year, ": ", e$message)
-  })
-}
 

### 2024-11-26 scripts/simulate_game_data.R

@@ -1,12 +1,10 @@
 #### Preamble ####
-# Purpose: Simulates and saves a dataset containing NFL performance data
-# for Tom Brady and Patrick Mahomes, retaining only specified columns.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Purpose: Simulates and saves a dataset containing NFL game data for Tom Brady and Patrick Mahomes
+# Author: Alexander Guarasci
+# Date: 26 November 2024
+# Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: The `tidyverse` and `arrow` packages must be installed.
-# Any other information needed? Make sure you are in the `starter_folder` rproj.
+# Pre-requisites: The `dplyr` and `arrow` packages must be installed.
 
 # Load necessary libraries
 library(dplyr)

### 2024-11-26 scripts/simulate_qbr_data.R

@@ -1,12 +1,10 @@
 #### Preamble ####
-# Purpose: Simulates and saves a dataset containing select QBR data for 
-# Tom Brady and Patrick Mahomes, retaining only specified columns.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
+# Purpose: Simulates and saves a dataset containing select QBR data for Tom Brady and Patrick Mahomes, retaining only specified columns.
+# Author: Alexander Guarasci
+# Date: 26 November 2024
+# Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: The `tidyverse` and `arrow` packages must be installed.
-# Any other information needed? Make sure you are in the `starter_folder` rproj.
+# Pre-requisites: The `dplyr` and `arrow` packages must be installed.
 
 # Load necessary libraries
 library(dplyr)

### 2024-11-26 scripts/test_analysis_qbr_data.R

@@ -1,3 +1,12 @@
+#### Preamble ####
+# Purpose: Tests analysis NFL QBR data for Tom Brady and Patrick Mahomes
+# Author: Alexander Guarasci
+# Date: 26 November 2024
+# Contact: alexander.guarasci@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: The `dplyr`, `arrow` and 'testthat' packages must be installed, and clean_data.R must be run
+
+
 # Load necessary libraries
 library(dplyr)
 library(arrow)

### 2024-11-26 scripts/test_game_analysis_data.R

@@ -1,3 +1,11 @@
+#### Preamble ####
+# Purpose: Tests analysis NFL game data for Tom Brady and Patrick Mahomes
+# Author: Alexander Guarasci
+# Date: 26 November 2024
+# Contact: alexander.guarasci@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: The `dplyr`, `arrow` and 'testthat' packages must be installed, and clean_data.R must be run
+
 library(testthat)
 library(arrow)
 library(dplyr)

### 2024-11-26 scripts/test_simulated_game_data.R

@@ -1,3 +1,12 @@
+#### Preamble ####
+# Purpose: Tests simulated NFL game data for Tom Brady and Patrick Mahomes
+# Author: Alexander Guarasci
+# Date: 26 November 2024
+# Contact: alexander.guarasci@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: The `dplyr`, `arrow` and 'testthat' packages must be installed, and simulate_game_data.R must be run
+
+
 # Load necessary libraries
 library(dplyr)
 library(arrow)

### 2024-11-26 scripts/test_simulated_qbr_data.R

@@ -1,3 +1,12 @@
+#### Preamble ####
+# Purpose: Tests simulated NFL QBR data for Tom Brady and Patrick Mahomes
+# Author: Alexander Guarasci
+# Date: 26 November 2024
+# Contact: alexander.guarasci@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: The `dplyr`, `arrow` and 'testthat' packages must be installed, and simulate_qbr_data.R must be run
+
+
 # Load necessary libraries
 library(dplyr)
 library(arrow)

### 2024-11-26 scripts/clean_data.R
preamble adjustment

@@ -4,7 +4,7 @@
 # Date: 26 November 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: The `nflverse` and `arrow` packages must be installed
+# Pre-requisites: The `nflverse` and `arrow` packages must be installed, run download_data.R 
 
 
 library(arrow)

### 2024-11-28 paper/paper.qmd
some minor adjustments to paper.qmd

@@
 
-The findings reveal that Brady’s legacy is unparalleled in its longevity and sheer accomplishment, while Mahomes’ career trajectory suggests he does have the potential to surpass Brady, it is highly unlikely that he does from a statistcal standpoint.
+The findings reveal that Brady’s legacy is unparalleled in its longevity and sheer accomplishment, while Mahomes’ career trajectory suggests he does have the potential to surpass Brady, it is highly unlikely that he does from a statistical standpoint.
 These insights are significant not only for appreciating their contributions to football but also for understanding the broader implications of what society values in its icons.
@@
 For example, receiving yards were excluded since quarterbacks rarely make receptions.
-The refined dataset highlights key metrics such as passing yards, completions, attempts, interceptions, sacks, sack fumbles (and lost), rushing yards, passing first downs, and a variety of quarterback rushing stats, which have become increasingly relevant in modern NFL play.
+The refined dataset highlights key metrics such as completions, attempts, passing yards, interceptions, fumbles, and a variety of quarterback rushing stats, which have become increasingly relevant in modern NFL play.
 
@@
 While alternative datasets could have been used, nflverse was chosen for its open accessibility, ease of integration with R, and the breadth of historical data.
-The inclusion of QBR data, although less granular, provided a complementary high-level view of player performance in both regular season and playoff contexts.
+The inclusion of QBR data, although less granular, provided a complimentary high-level view of player performance in both regular season and playoff contexts.
 
@@
 Most of the variables used in this analysis are relatively straight forward.
-Passing yards, is the vertical distance a passing play gains, so for example, if a quarterback throws a pass, the reciever catches 10 yards down the field and runs for another 10 yards before being tackled or going out of bounds, that will count as 20 passing yards for the Quarterback.
-Rushing yards are just how many yards a player runs with ball past the line of scrimmage, if the ball is caught as a pass, it does not count towards rushing yards.
+Passing yards is the vertical distance a passing play gains.
+For example, if a quarterback throws a pass and the receiver catches 10 yards down the field and runs for another 10 yards before being tackled or going out of bounds, that will count as 20 passing yards for the Quarterback.
+Rushing yards are the number of yards a player runs with the ball past the line of scrimmage - if the ball is caught as a pass, it does not count towards rushing yards.
 Passing touchdowns are simply a passing play that turns into a touchdown.
-Interceptions are just passes made by a quarterback and caught by the other team.
+Interceptions are passes made by a quarterback and caught by the other team.
 
@@
 
-Although the previous graphs were useful in getting a picture of the two players respective careers, what makes a player great in any sport is how they perform in the clutch.
+Although the previous graphs were useful in getting a picture of the two players' respective careers, what makes a player great in any sport is how they perform in a clutch situation.
 In other words, big players make big plays.
@@
 
-[@tbl-brady-mahomes-career-game-stats], shows the per game projected averages for Mahomes and compares it with Bradys stats.
+[@tbl-brady-mahomes-career-game-stats], shows the per game projected averages for Mahomes and compares it with Brady's stats.
 
@@
 ## What is done in this paper? 
-This paper provides a rigorous comparison of Tom Brady and Patrick Mahomes to evaluate whether Mahomes could eventually surpass Brady as the greatest quarterback of all time (GOAT).
+This paper provides a rigorous statistical comparison of Tom Brady and Patrick Mahomes to evaluate whether Mahomes could eventually surpass Brady as the greatest quarterback of all time (GOAT).
 Through comprehensive data analysis, the study examines key performance metrics across both players’ careers, incorporating traditional and advanced statistics, including ESPN’s Total Quarterback Rating (QBR).
@@
 One significant insight from this paper is the impact of career longevity on perceptions of greatness.
-Tom Brady’s career, spanning over two decades, showcases not only impressive cumulative statistics but also a sustained level of excellence, particularly in high-stakes postseason environments.
+Tom Brady’s career, spanning over two decades, showcases not only impressive cumulative statistics but also a sustained level of excellence, particularly in high-stakes post-season environments.
 This longevity has set an unprecedented benchmark for future players.
@@
 Other than that, the only thing left to do is wait and see.
-The Mahomes led Cheifs are 9-1 as this is being written.
-They are coming off back to back Superbowl victories, they are in the midst of a dynasty and all that is left to do is witness greatness unfold.
+The Mahomes led Chiefs are 9-1 as this is being written.
+They are coming off back-to-back Super Bowl victories, they are in the midst of a dynasty and all that is left to do is witness greatness unfold.
 

### 2024-11-30 paper/paper.qmd
minor change to paper
some more edits to paper

@@
 date-format: long
-abstract: "This paper compares NFL quarterbacks Tom Brady and Patrick Mahomes, focusing on their statistical performances.
-Analyzing their respective career data, this paper uses predictive modelling to forecast Patrick Mahomes future.
+abstract: "This paper compares National Football League (NFL) quarterbacks Tom Brady and Patrick Mahomes, focusing on their statistical performances.
+Analyzing their respective career data, this paper uses predictive modelling to forecast Patrick Mahomes’ future.
 It finds that while both have impressive career statistics, ultimately, playoff victories, particularly Super Bowl wins, serve as a more definitive measure of greatness.
-By examining these players, the paper highlights how the NFL’s dynamic context impacts player evaluation and legacy assessment."
+By examining the performance of these players, the paper highlights how the NFL’s dynamic context impacts player evaluation and legacy assessment."
 format: 
@@
 To answer this question, this analysis draws on statistical data, examining past performance, and broader career trajectories.
-It evaluates what each quarterback has achieved to date and aims to project Patrick Mahomes future.
+It evaluates what each quarterback has achieved to date and aims to project Patrick Mahomes’ future.
 
-The findings reveal that Brady’s legacy is unparalleled in its longevity and sheer accomplishment, while Mahomes’ career trajectory suggests he does have the potential to surpass Brady, it is highly unlikely that he does from a statistical standpoint.
+The findings reveal that Brady’s legacy is unparalleled in its longevity and sheer accomplishment.
+While Mahomes’ career trajectory suggests he does have the potential to surpass Brady, it is highly unlikely that he does from a statistical standpoint.
 These insights are significant not only for appreciating their contributions to football but also for understanding the broader implications of what society values in its icons.
@@
 The paper is structured as follows: the first section provides an explanation of the data used in this paper, what it shows and where it comes from.
-The second section discusses the model used to predict the future of Patrick Mahomes career.
+The second section discusses the model used to predict the future of Patrick Mahomes’ career.
 The third section shows the results of the model.
-The final section discusses the implication of the model as well as provides more insight into the two players legacies.
+The final section discusses the implication of the model as well as provides more insight into the two players’ legacies.
 
@@
 
-The data used in this analysis originates from nflverse [@citenflverse], a comprehensive collection of packages tailored for National Football League (NFL) data.
-These packages aggregate statistics, player information, and game results spanning several decades.
+The data used in this analysis originates from nflverse [@citenflverse], a comprehensive collection of packages tailored for NFL data.
+These packages aggregate statistics, player information and game results spanning several decades.
 For this paper, the focus was narrowed to quarterback (QB) performance data, specifically for Tom Brady and Patrick Mahomes.
@@
 Brady’s consistency and durability throughout his career, as seen in his passing and rushing stats, have firmly established him as one of the greatest quarterbacks in NFL history.
-Mahomes, on the other hand, shows great potential and is already a future first balot Hall of Famer, but he has yet to match Brady’s career volume.
+Mahomes, on the other hand, shows great potential and is already a future first ballot Hall of Famer, but he has yet to match Brady’s career volume.
 

### 2024-12-01 paper/paper.qmd
Update paper.qmd

@@
 
-While statistical achievements and accolades are critical, they alone cannot encapsulate the essence of greatness.
+While statistical achievements and accolades are important, they alone cannot encapsulate the essence of greatness.
 Both Brady and Mahomes boast extraordinary numbers, yet their legacies diverge in ways that reflect deeper, qualitative factors such as resilience, leadership, and the ability to deliver in high-pressure moments.
@@
 
-The findings reveal that Brady’s legacy is unparalleled in its longevity and sheer accomplishment.
+The findings reveal that Brady’s legacy is unequaled in its longevity and sheer accomplishment.
 While Mahomes’ career trajectory suggests he does have the potential to surpass Brady, it is highly unlikely that he does from a statistical standpoint.
@@
 
-The data used in this analysis originates from nflverse [@citenflverse], a comprehensive collection of packages tailored for NFL data.
+The data used in this analysis originates from nflverse [@citenflverse], a collection of packages tailored for NFL data.
 These packages aggregate statistics, player information and game results spanning several decades.
@@
 The primary dataset included QB-specific data from every NFL game played since 2000, encompassing over 55 columns detailing various player performance metrics.
-Additional data sources were used for advanced metrics, notably ESPN’s Quarterback Rating (QBR), which encapsulates QB efficiency across multiple dimensions of gameplay.
+Additional data sources were used for ESPN’s Quarterback Rating (QBR), which encapsulates QB efficiency across multiple dimensions of gameplay.
 
 The analysis focuses on comparing the careers of two of the greatest football players of all time: Tom Brady and Patrick Mahomes.
-Both players’ careers are comprehensively represented in the dataset, allowing for an in-depth comparison of Mahomes’ trajectory against Brady’s completed career.
+Both players’ careers are completely represented in the dataset, allowing for an in-depth comparison of Mahomes’ trajectory against Brady’s completed career.
 To achieve this, the dataset was filtered to include only these two players, with irrelevant data removed.
@@
 
-Additionally, a second dataset containing ESPN’s advanced Quarterback Rating (QBR) metric was incorporated to analyze playoff performance across their careers.
+Additionally, a second dataset containing ESPN’s Quarterback Rating (QBR) metric was incorporated to analyze playoff performance across their careers.
 This dataset, also filtered to focus on Brady and Mahomes, provided season and postseason averages, though it lacked the granularity of the primary dataset.
@@
 ## Why This Dataset?
-This dataset was selected for its granularity and comprehensive scope, capturing both raw and advanced metrics for NFL quarterbacks.
+This dataset was selected for its granularity and scope, capturing raw metrics for NFL quarterbacks.
 While alternative datasets could have been used, nflverse was chosen for its open accessibility, ease of integration with R, and the breadth of historical data.
@@
 
-Another critical element of football data collection lies in the placement of the ball after a play.
+Another important element of football data collection lies in the placement of the ball after a play.
 Typically, a quarterback throws the ball, it is caught by a receiver, and the play concludes when the receiver either goes out of bounds or is tackled.
@@
 While this measurement is essential for maintaining game flow and accuracy, it relies heavily on visual judgment and is thus more art than science.
-Although teams may challenge the spot in crucial situations, most ball placements remain subjective, resulting in potential variability in the dataset.
+Although teams may challenge the spot in pivotal situations, most ball placements remain subjective, resulting in potential variability in the dataset.
 
 These nuances highlight the challenges inherent in translating the dynamic, real-time events of a football game into static data entries.
-The dataset used in this analysis reflects not only the physical phenomena observed on the field but also the interpretive decisions made by officials, which can influence critical measures like completions and ball placement.
+The dataset used in this analysis reflects not only the physical phenomena observed on the field but also the interpretive decisions made by officials, which can influence measures like completions and ball placement.
 This underscores the importance of considering the limitations and subjectivity embedded within the data when drawing conclusions about player performance and game outcomes.
@@
 The most nuanced variable used in the analysis is ESPN QBR.
-This is a stat that aims to be all encompassing with regard to how a quarterback is playing.
+This is a stat that aims explain the totality of how a quarterback is playing.
 It is a scale from 0-100 with 50 being average.
@@
 ```
-[@fig-total-rushing-yards] reveals an interesting contrast in rushing yards between the two quarterbacks.
+[@fig-total-rushing-yards] shows an interesting contrast in rushing yards between the two quarterbacks.
 Despite playing over three times as many games as Mahomes, Tom Brady falls significantly short, with Mahomes amassing more than twice as many rushing yards.
@@
 [@fig-interceptions-per-season] highlights the number of interceptions thrown by each player per season.
-Interceptions are critical errors that often stem from poor decision-making or miscommunication.
+Interceptions are significant errors that often stem from poor decision-making or miscommunication.
 This graph provides insight into the players’ decision-making abilities and their capacity to minimize costly turnovers.
@@
 We designed a predictive model to estimate Patrick Mahomes' cumulative career statistics over an additional 200 games.
-The model leverages historical game data from Tom Brady as a benchmark and Patrick Mahomes' existing career data to generate predictions for key performance metrics.
+The model is trained on historical game data from Tom Brady as a benchmark and Patrick Mahomes' existing career data to generate predictions for key performance metrics.
 
@@
    - `carries`, `rushing_yards`, `rushing_tds`.  
-   - While not the primary focus, rushing performance is crucial for dual-threat quarterbacks like Mahomes.
+   - While not the primary focus, rushing performance is an insightful metric for dual-threat quarterbacks like Mahomes.
 
@@
    - `sack_fumbles`.  
-   - Turnovers are critical in evaluating overall impact.
+   - Turnovers are highly important in evaluating overall impact.
 
@@
 Brady also has more interceptions (252 compared to Mahomes’ 213.59).
-These differences underscore Brady’s long career and high volume of passing, which has contributed to his unparalleled career achievements.
+These differences underscore Brady’s long career and high volume of passing, which has contributed to his unrivaled career achievements.
 
@@
 
-The analysis of per-game statistics reveals a clear edge for Patrick Mahomes over Tom Brady across nearly every metric.
+The analysis of per-game statistics shows a clear edge for Patrick Mahomes over Tom Brady across nearly every metric.
 Mahomes completes slightly more passes per game (23.94) than Brady (23.50), despite attempting a comparable number of throws (36.74 for Mahomes vs.
@@
 This paper provides a rigorous statistical comparison of Tom Brady and Patrick Mahomes to evaluate whether Mahomes could eventually surpass Brady as the greatest quarterback of all time (GOAT).
-Through comprehensive data analysis, the study examines key performance metrics across both players’ careers, incorporating traditional and advanced statistics, including ESPN’s Total Quarterback Rating (QBR).
+Through comprehensive data analysis, the study examines key performance metrics across both players’ careers, incorporating traditional and more high-level statistics, including ESPN’s Total Quarterback Rating (QBR).
 Additionally, the paper develops a predictive model to estimate Mahomes’ career trajectory if he plays 200 more games.
@@
 While both players have put up astonishing numbers throughout their careers, attempting to compare them purely through stats is an oversimplification of their legacies.
-Statistics, while important, fail to account for the intangibles—the ability to perform under pressure, the knack for leading a team to victory in critical moments, and the sheer will to win when it matters most.
+Statistics, while important, fail to account for the intangibles—the ability to perform under pressure, the knack for leading a team to victory in clutch moments, and the sheer will to win when it matters most.
 
 When evaluating greatness, Super Bowl victories tend to stand as the ultimate benchmark, and here lies the defining gap between Brady and Mahomes.
-Tom Brady has set an unparalleled standard with seven Super Bowl wins, a number that remains a daunting goal for any quarterback to reach.
+Tom Brady has set an unprecedented standard with seven Super Bowl wins, a number that remains a daunting goal for any quarterback to reach.
 Mahomes, as brilliant as he has been in his career so far, still has a long way to go in this regard.
@@
 
-This also underscores a critical point about Mahomes: his ability to consistently elevate his game when the stakes are highest.
+This also highlights another point about Mahomes: his ability to consistently elevate his game when the stakes are highest.
 Winning playoff games is no small feat, yet Mahomes has made it look routine, guiding his team deep into the postseason year after year.
@@
 ## What are some weaknesses of what was done?
-While this paper offers valuable insights, there are several limitations worth acknowledging.
+While this paper offers some insights, there are several limitations worth acknowledging.
 First, the predictive model relies on the assumption that Mahomes’ career trajectory will follow trends observed in Brady’s career, which may not account for unforeseen injuries, team dynamics, or changes in the NFL landscape.
@@
 Advances in machine learning could also enable the development of more sophisticated models to predict player performance under varying conditions.
-Lastly, further exploration into how the definition of greatness evolves across sports can provide context for comparing athletes from different eras, enhancing the broader conversation about legacy and excellence.
+Lastly, looking into how the definition of greatness evolves across sports can provide context for comparing athletes from different eras, enhancing the broader conversation about legacy and excellence.
 
@@
 The NFLverse dataset depends on a combination of human and technological data collection methods.
-Human referees make critical decisions on plays, introducing subjectivity into measurements such as completions and ball placement:
+Human referees make decisions on plays, introducing subjectivity into measurements such as completions and ball placement:
 
@@
 
-Building on these frameworks, this analysis incorporated advanced metrics like ESPN's QBR, which contextualizes a quarterback's contributions within the broader game environment.
+Building on these frameworks, this analysis incorporated metrics like ESPN's QBR, which contextualizes a quarterback's contributions within the broader game environment.
 While QBR addresses some limitations of traditional statistics, it remains an observational metric and is thus subject to the same biases and variability.
@@
 The NFLverse dataset represents one of the most comprehensive sources of quarterback performance data available, but its observational nature necessitates careful consideration of biases, subjectivity, and measurement variability.
-By employing normalization techniques, incorporating advanced metrics, and referencing existing literature, this analysis seeks to provide a balanced and nuanced comparison of Brady and Mahomes.
+By employing normalization techniques, and referencing existing literature, this analysis seeks to provide a balanced and nuanced comparison of Brady and Mahomes.
 
-While observational data cannot establish causality, it offers valuable insights into patterns and trends, particularly when analyzed with an awareness of its limitations.
+While observational data cannot establish causality, it offers insights into patterns and trends, particularly when analyzed with an awareness of its limitations.
 Future work could enhance these findings by incorporating experimental methods, such as simulations or controlled studies, to isolate causal factors influencing quarterback performance.

### 2024-12-01 paper/paper.qmd
Update paper.qmd

@@
 title: "Defining Greatness"
-subtitle: "A Comparative Analysis of Tom Brady and Patrick Mahomes"
+subtitle: "Mahomes’ Path to Immortality vs. Brady’s Unmatched Legacy"
 author: 
@@
 
-The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Arrow [@citearrow], Knitr [@citeknitr], Testthat [@citett] packages.
+The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Arrow [@citearrow], Knitr [@citeknitr], and Testthat [@citett] packages.
 The primary dataset included QB-specific data from every NFL game played since 2000, encompassing over 55 columns detailing various player performance metrics.
@@
 Rushing yards are the number of yards a player runs with the ball past the line of scrimmage - if the ball is caught as a pass, it does not count towards rushing yards.
-Passing touchdowns are simply a passing play that turns into a touchdown.
+Passing touchdowns are simply a passing play that turns into a touchdown and rushing touchdowns are a rush that turns into a touchdown.
 Interceptions are passes made by a quarterback and caught by the other team.
 
-The most nuanced variable used in the analysis is ESPN QBR.
+The most complicated variable used in the analysis is ESPN QBR.
 This is a stat that aims explain the totality of how a quarterback is playing.
@@
 ESPN says "it incorporates all of a quarterback’s contributions to winning, including how he impacts the game on passes, rushes, turnovers and penalties.
-Also, since QBR is built from the play level, it accounts for a team’s level of success or failure on every play to provide the proper context and then allocates credit to the quarterback and his teammate to produce a clearer measure of quarterback efficiency." Overall, it aims to give an overview of how well a quarterback played.
+Also, since QBR is built from the play level, it accounts for a team’s level of success or failure on every play to provide the proper context and then allocates credit to the quarterback and his teammate to produce a clearer measure of quarterback efficiency" [@citeespn].
+Overall, it aims to give an overview of how well a quarterback played.
 
@@
 This extraordinary season set the stage for Mahomes’ rise as one of the game’s best quarterbacks, though it ultimately ended in a playoff loss to Brady.
-The graph underscores how both players have reached unprecedented heights in their respective careers.
+The graph underscores how both players have reached incredible heights in their respective careers.
 ```{r}
@@
 [@fig-passing-yard-distribution] examines the distribution of passing yards per game for each player.
-While both quarterbacks exhibit similar averages, Brady’s larger sample size, due to his longer career, leads to greater variability in his distribution.
-This graph captures the consistency of both players but also highlights the statistical advantage that longevity brings to career totals.
-Mahomes’ tighter distribution reflects his efficiency and performance within a shorter time frame.
+While both quarterbacks exhibit similar averages, Brady’s larger sample size, due to his longer career, shows higher highs and lower lows.
+This graph captures the consistency of both players but also highlights the significance of longevity with respect to great and poor performances.
+Mahomes’ distribution reflects his efficiency and performance within a shorter time frame.
 
@@
 Interceptions are significant errors that often stem from poor decision-making or miscommunication.
-This graph provides insight into the players’ decision-making abilities and their capacity to minimize costly turnovers.
+This graph allows us to see the players’ decision-making abilities and their capacity to minimize costly turnovers.
 ```{r}
@@
 ## What is something that we learn about the world? 
-One significant insight from this paper is the impact of career longevity on perceptions of greatness.
+One significant thing that we learn from this paper is the impact of career longevity on perceptions of greatness.
 Tom Brady’s career, spanning over two decades, showcases not only impressive cumulative statistics but also a sustained level of excellence, particularly in high-stakes post-season environments.
@@
 Greatness in sports, especially in the context of legendary quarterbacks like Tom Brady and Patrick Mahomes, cannot be boiled down to statistics alone.
-While both players have put up astonishing numbers throughout their careers, attempting to compare them purely through stats is an oversimplification of their legacies.
+While both players have put up astonishing numbers throughout their careers, attempting to compare them purely through statistics is an oversimplification of their legacies.
 Statistics, while important, fail to account for the intangibles—the ability to perform under pressure, the knack for leading a team to victory in clutch moments, and the sheer will to win when it matters most.
@@
 But Mahomes’ trajectory, both statistically and in terms of team success, suggests that he could one day close the gap.
-What we learn about the world through their stories is that greatness is not about one-dimensional measures like stats—it’s about how athletes respond to challenges, seize opportunities, and etch their names in history through consistent excellence and defining moments.
+What we learn about the world through their stories is that greatness is not about one-dimensional measures like statistics—it’s about how athletes respond to challenges, seize opportunities, and etch their names in history through consistent excellence and defining moments.
 
@@
 Second, the analysis predominantly focuses on individual statistics without fully considering the impact of team success, coaching, or supporting cast, all of which significantly contribute to a quarterback’s achievements.
-Lastly, the reliance on ESPN’s QBR as a key metric introduces subjectivity, as its proprietary formula may not fully capture the nuances of on-field performance.
+Lastly, ESPN’s QBR as a metric introduces subjectivity, as its proprietary formula may not fully capture the nuances of on-field performance.
 
@@
 Observational data inherently captures events as they unfold in their natural context.
-In the case of the NFLverse dataset, this includes game statistics, player-specific metrics, and play-by-play details recorded during live games.
+In the case of the nflverse dataset, this includes game statistics, player-specific metrics, and play-by-play details recorded during live games.
 Unlike experimental data, there is no controlled randomization or intervention, meaning that any observed relationships are correlational rather than causal.
@@
 
-Despite its observational nature, the NFLverse dataset is highly granular, capturing over 55 variables per game, which mitigates some biases by enabling a comprehensive multivariate analysis.
+Despite its observational nature, the nflverse dataset is highly granular, capturing over 55 variables per game, which mitigates some biases by enabling a comprehensive multivariate analysis.
 However, the dataset remains subject to inherent challenges, including subjective measurements and incomplete data points.
@@
 
-To address these biases, the analysis normalized certain statistics, such as average passing yards per game, to account for differences in season length and rule changes.
+To address these biases, the analysis normalized certain statistics, showing statistical averages per game, to account for differences in season length.
 Additionally, Mahomes' projected career totals were included to balance the comparison with Brady's completed career.
@@
 
-To assess the impact of these subjective measurements, a simulation was performed on hypothetical data with varying degrees of error in ball placement (e.g., ±1 yard, ±2 yards).
-The results showed that small measurement errors had a negligible effect on aggregate statistics, such as total passing yards over a season.
-However, these discrepancies could influence specific game outcomes or play-by-play analyses.
-
-## Linking to Literature: Observational Data and Sport Analytics
-
-The challenges outlined above align with findings in the sports analytics literature.
-Researchers such as Goldsberry (2019) have highlighted the limitations of observational data in basketball, emphasizing that player performance is deeply intertwined with contextual variables.
-Similarly, Burke et al.
-(2020) demonstrated the importance of adjusting football statistics for external factors, such as opponent strength and game context, when evaluating player performance.
-
 Building on these frameworks, this analysis incorporated metrics like ESPN's QBR, which contextualizes a quarterback's contributions within the broader game environment.
@@
 
-The NFLverse dataset represents one of the most comprehensive sources of quarterback performance data available, but its observational nature necessitates careful consideration of biases, subjectivity, and measurement variability.
-By employing normalization techniques, and referencing existing literature, this analysis seeks to provide a balanced and nuanced comparison of Brady and Mahomes.
+The nflverse dataset represents one of the most comprehensive sources of quarterback performance data available, but its observational nature necessitates careful consideration of biases, subjectivity, and measurement variability.
+Ultimately this analysis seeks to provide a balanced comparison of Brady and Mahomes.
 

### 2024-12-01 paper/paper.qmd
Update paper.qmd

@@
 For example, receiving yards were excluded since quarterbacks rarely make receptions.
-The refined dataset highlights key metrics such as completions, attempts, passing yards, interceptions, fumbles, and a variety of quarterback rushing stats, which have become increasingly relevant in modern NFL play.
+The refined dataset highlights key metrics such as completions, attempts, passing yards, interceptions, and a variety of quarterback rushing stats, which have become increasingly relevant in modern NFL play.
 
@@
 1. **Passing Metrics**:  
-   - `completions`, `attempts`, `passing_yards`, `passing_tds`, `interceptions`, `passing_air_yards`.  
+   - `completions`, `attempts`, `passing_yards`, `passing_tds`, `interceptions`.  
    - These metrics directly influence a quarterback’s overall contribution to team success.
@@
    - While not the primary focus, rushing performance is an insightful metric for dual-threat quarterbacks like Mahomes.
-
-3. **Ball Security**:  
-   - `sack_fumbles`.  
-   - Turnovers are highly important in evaluating overall impact.
 
@@
 columns_to_sum <- c(
-  "completions", "attempts", "passing_yards", "passing_tds", "interceptions",
-  "sack_fumbles", "passing_air_yards", "carries", "rushing_yards", "rushing_tds"
+  "completions", "attempts", "passing_yards", "passing_tds", "interceptions", "carries", "rushing_yards", "rushing_tds"
 )
@@
     `INTs` = interceptions,
-    `Fumbles` = sack_fumbles,
-    `Air Yds` = passing_air_yards,
     `Carries` = carries,
@@
 # Generate the table using tinytable
-tt(final_table, width = 1.5) |> style_tt( fontsize = .5)
+tt(final_table, width = 1.5) |> style_tt( fontsize = 1)
 ```
@@
 In terms of scoring, Mahomes demonstrates a superior passing touchdown rate, averaging 2.04 touchdowns per game compared to Brady’s 1.94.
-Both players maintain an impressive level of ball security, with nearly identical interception rates per game (0.66), but Mahomes edges Brady in minimizing sack-related turnovers, with 0.27 sack fumbles per game versus Brady’s 0.30.
-This reflects Mahomes’ adaptability under pressure and his efficiency in avoiding costly mistakes.
+Both players maintain an impressive level of ball security, with nearly identical interception rates per game (0.66).
 

### 2024-12-01 paper/model_card.qmd
Add files via upload

@@
 - Dataset: Player-level NFL game statistics from `merged_player_game.parquet`.  
-- Features: 16 numerical features, including completions, attempts, passing yards, passing touchdowns, interceptions, and rushing statistics.  
+- Features: Includes completions, attempts, passing yards, passing touchdowns, interceptions, and rushing statistics.  
 - Training Data: Filtered observations of Tom Brady's career game statistics.  
@@
 
-**Evaluation Metrics**: The model's fit and prediction accuracy were evaluated using the following metrics:  
-- Adjusted R² values for multivariate regression.  
-- Mean squared error (MSE) for the fitted model.
+**Evaluation Metrics**: The model's fit and prediction accuracy were evaluated using RMSE following
 
 **Performance on Training Data (Tom Brady)**:  
-- The model successfully captured trends in Tom Brady’s career statistics, showing high adjusted R² values for most features.  
-- No missing data points were included in the training subset.
-
-**Projection Validation**:  
-- Model predictions were cross-referenced with Mahomes' historical averages.  
-- Assumes statistical continuity in performance.
+- The model somewhat successfully captured trends in Tom Brady’s career statistics
 
@@
 1.
-**Contextual Limitations**: The model assumes that Patrick Mahomes' future performance mirrors his historical averages and does not account for potential changes in playing style, team composition, or career disruptions.
+**Contextual Limitations**: The model assumes that Patrick Mahomes' future performance mirrors Tom Brady's and does not account for potential changes in playing style, team composition, or career disruptions.
 

### 2024-12-01 paper/paper.qmd
Add files via upload

@@
 
-The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Arrow [@citearrow], Knitr [@citeknitr], and Testthat [@citett] packages.
+The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Caret [@citecaret], Arrow [@citearrow], Knitr [@citeknitr], and Testthat [@citett] packages.
 The primary dataset included QB-specific data from every NFL game played since 2000, encompassing over 55 columns detailing various player performance metrics.
@@
 Interceptions are passes made by a quarterback and caught by the other team.
+Carries are how many rush attempts the QB has.
 
@@
 
-## Model Overview
+## Overview
 
-We designed a predictive model to estimate Patrick Mahomes' cumulative career statistics over an additional 200 games.
-The model is trained on historical game data from Tom Brady as a benchmark and Patrick Mahomes' existing career data to generate predictions for key performance metrics.
+The goal of this analysis is to predict Patrick Mahomes' lifetime career statistics over 200 additional games using a regression model.
+This model uses data from Tom Brady's career as a proxy for player performance and is designed to forecast Mahomes' future performance across key metrics, such as completions, attempts, passing yards, rushing yards, and touchdowns.
 
-### Mathematical Notation
+## Data
 
-The model is a multivariate linear regression defined as:
+The dataset used in this analysis is a combination of game-level statistics for two players: Tom Brady and Patrick Mahomes.
+The data includes multiple variables that track each player's performance across different games.
+The key features used in this model are:
+
+- **completions**: The number of successful passes thrown.
+- **attempts**: The total number of pass attempts.
+- **passing_yards**: The total yards gained from passing attempts.
+- **passing_tds**: The total number of passing touchdowns.
+- **interceptions**: The total number of interceptions thrown.
+- **carries**: The number of rushing attempts.
+- **rushing_yards**: The total rushing yards gained.
+- **rushing_tds**: The total number of rushing touchdowns.
+
+The model used in this analysis is a multiple linear regression model.
+We choose linear regression because it is a simple yet effective approach for modeling continuous outcomes and provides easy interpretability.
+
+### Equation for the Model
+
+The model predicts the following variables for Patrick Mahomes:
+
+- \( \text{completions}, \text{attempts}, \text{passing yards}, \text{passing tds}, \text{interceptions}, \text{carries}, \text{rushing yards}, \text{rushing tds} \)
+
+The general form of the multiple linear regression model is:
 
 $$
-\hat{y}_i = \beta_0 + \sum_{j=1}^p \beta_j x_{ij} + \epsilon_i
+Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_n X_n + \epsilon
 $$
@@
 
-- $\hat{y}_i$: Predicted value of the response variable (e.g., total passing yards).  
-- $\beta_0$: Intercept term.  
-- $\beta_j$: Coefficients for predictors $x_{ij}$, representing the influence of the $j$-th feature on the response variable.  
-- $x_{ij}$: Observed value of the $j$-th predictor for the $i$-th observation.  
-- $\epsilon_i$: Residual error term, assumed to be normally distributed with mean 0.
+- $Y$ represents the outcome variable (e.g., completions, attempts, passing yards, etc.).
+- $X_1, X_2, \dots, X_n$ are the predictor variables (game statistics such as completions, attempts, etc.).
+- $\beta_0$ is the intercept.
+- $\beta_1, \beta_2, \dots, \beta_n$ are the coefficients for each predictor.
+- $\epsilon$ is the error term.
 
-We estimate $\beta_0$ and $\beta_j$ using Ordinary Least Squares (OLS) on Tom Brady's historical data, as his career provides a well-documented trajectory for long-term performance.
+### Training and Validation
 
-### Variables and Justification
+We trained the model using data from Tom Brady's career, specifically using the following steps:
 
-The predictors ($x_j$) included in the model are chosen based on their relevance to a quarterback's performance:
+1. **Data Preparation**: We selected relevant features and ensured there were no missing values in the training dataset.
+2. **Model Creation**: A multiple linear regression model was created using the `lm()` function in R. The model predicts the selected statistics for each game.
+3. **Training/Test Split**: The dataset was split into training (80%) and testing (20%) sets to evaluate model performance.
+4. **Prediction**: The trained model was used to predict Mahomes' statistics for the next 200 games.
 
-1. **Passing Metrics**:  
-   - `completions`, `attempts`, `passing_yards`, `passing_tds`, `interceptions`.  
-   - These metrics directly influence a quarterback’s overall contribution to team success.
+We also calculated the Root Mean Squared Error (RMSE) for each predicted variable in the test set as an evaluation metric.
+RMSE is a commonly used metric to measure the accuracy of regression models, and it is calculated as:
 
-2. **Rushing Metrics**:  
-   - `carries`, `rushing_yards`, `rushing_tds`.  
-   - While not the primary focus, rushing performance is an insightful metric for dual-threat quarterbacks like Mahomes.
+$$
+RMSE = \sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2}
+$$
 
-These features ensure the model captures both primary and auxiliary aspects of quarterback performance.
+Where:
 
-### Model Assumptions
+- $y_i$ is the actual value of the statistic.
 
-1. **Linearity**: The relationship between predictors and response variables is linear.  
-2. **Independence**: Residuals are independent across observations.  
-3. **Homoscedasticity**: Residuals have constant variance.  
-4. **Normality**: Residuals are normally distributed.  
-5. **Stationarity**: Mahomes' performance trajectory will follow trends observed in Brady's career.
+- $\hat{y}_i$ is the predicted value.
 
-### Software and Implementation
+- $n$ is the number of test samples.
 
-The model was implemented using R, leveraging the `lm` function from the `stats` package for regression.
-The dataset was preprocessed using `tidyverse`, and predictions were calculated for an additional 200 games using the coefficients estimated from Brady's data.
+The RMSE values for each variable in the test set are as follows:
 
-### Validation and Diagnostics
+completions: 6.01,
+attempts: 9.57,
+passing yards: 75.90,
+passing touchdowns: 1.28,
+interceptions: 0.80,
+carries: 1.72,
+rushing yards: 4.70 and
+rushing touchdowns: 0.15
 
-1. **Train-Test Split**: The Brady dataset was split into training (80%) and testing (20%) sets. The model was trained on the training set and validated on the testing set.  
-2. **Error Metrics**:  
-   - **Root Mean Squared Error (RMSE)**: Measures the model's predictive accuracy.  
-   - **Mean Absolute Error (MAE)**: Evaluates average prediction error.  
-3. **Residual Analysis**:  
-   - Residual plots confirmed the assumptions of homoscedasticity and normality.  
-4. **Out-of-Sample Testing**: Predicted Mahomes' statistics on his observed data and compared with actual results to ensure alignment.
+These RMSE values indicate that the model has some level of predictive accuracy but also considerable error in certain variables, such as passing yards.
+The RMSE is higher than expected, which can be attributed to the limited dataset size used in this model.
+The dataset contains a relatively small number of games, which leads to higher variability and uncertainty in the predictions.
 
@@
    - Strengths: Captures non-linear relationships.  
-   - Weaknesses: Tends to overfit without pruning; less interpretable than linear regression.
+   - Weaknesses: Tends to overfit without pruning; less interpretable than linear regression, and still had high error.
 
@@
    - Strengths: Allows incorporation of priors, producing probabilistic predictions.  
-   - Weaknesses: Increased complexity and computational requirements.
+   - Weaknesses: Increased complexity and computational requirements and also had high error.
 
@@
    - Linear regression was chosen for its balance of simplicity, interpretability, and performance.
-It aligns with the assumption that a quarterback’s performance trends over time can be captured linearly.
 
-### Limitations
+## Limitiations
 
-1. **Career Longevity**:  
-   - The model assumes Mahomes will continue performing at a consistent level for 200 games.
-This may not account for potential injuries, performance decline, or external factors.
-This is already a very generous assumption, considering only two quarterbacks in NFL history have played more than 324 career games, but because Brady has played 381 games, it is important for our analysis to assume that Mahomes will be able to sustain success.
-2. **Sample Bias**:  
-   - Relying on Brady's career as a benchmark may introduce bias, as it assumes Mahomes will follow a similar trajectory.  
-3. **Feature Engineering**:  
-   - Excluding contextual factors like team strength or play style may limit predictive accuracy.
+While the RMSE values suggest some error in the predictions, this is expected due to the limited amount of available data.
+The performance of the model could likely improve with a larger and more balanced dataset that includes a broader range of game conditions.
+Additionally, while we used Tom Brady’s statistics as a proxy for Mahomes, differences in their playing styles and team dynamics could also contribute to the prediction error.
 
-### Conclusion
+The linear regression model assumes that the relationships between the predictors and the outcomes are linear, which may not always be the case in real-world scenarios.
+Further exploration of alternative models, such as non-linear regression or machine learning techniques like decision trees, could be considered for improving predictions.
 
-The linear regression model provides a robust yet interpretable framework for predicting Mahomes' lifetime stats.
-It is grounded in historical data and validated through residual analysis and error metrics.
-While limitations exist, the model effectively captures the essence of Mahomes' expected performance trajectory based on Brady's historical patterns.
+## Conclusion
+
+This model provides a reasonable estimate of Patrick Mahomes’ lifetime career statistics over 200 additional games based on Tom Brady’s historical data.
+Although the RMSE values suggest room for improvement, the model serves as a useful starting point for projecting Mahomes’ future performance.
+Future work could include expanding the dataset, exploring non-linear modeling techniques, and accounting for player-specific factors that might differ between Brady and Mahomes.
 
@@
  [@tbl-brady-mahomes-career-stats] compares Tom Brady’s career statistics with Patrick Mahomes’ projected lifetime statistics.
-Tom Brady, with 8,954 completions and 13,973 attempts, has significantly more career completions and attempts than Mahomes, who is projected to finish with 7,756.72 completions and 11,902.61 attempts.
-Brady’s total passing yards of 102,616 far exceed Mahomes’ projected 89,548.38, highlighting Brady’s longevity and consistent performance.
-Similarly, Brady has thrown 738 passing touchdowns, whereas Mahomes is projected to finish with 660.38.
-Brady also has more interceptions (252 compared to Mahomes’ 213.59).
+Tom Brady, with 8,954 completions and 13,973 attempts, has significantly more career completions and attempts than Mahomes, who is projected to finish with 7,763 completions and 11,925 attempts.
+Brady’s total passing yards of 102,616 far exceed Mahomes’ projected 89,160, highlighting Brady’s longevity and consistent performance.
+Similarly, Brady has thrown 738 passing touchdowns, whereas Mahomes is projected to finish with 654.
+Brady also has more interceptions (252 compared to Mahomes’ 216).
 These differences underscore Brady’s long career and high volume of passing, which has contributed to his unrivaled career achievements.
@@
 When comparing rushing statistics, Mahomes stands out as the more mobile quarterback.
-While Brady’s rushing stats are modest, with 1,256 rushing yards and 35 rushing touchdowns, Mahomes is projected to amass 3,282.88 rushing yards and 36.28 rushing touchdowns, showing a clear advantage in rushing ability.
-Additionally, Mahomes is expected to have 319.90 rushing first downs, surpassing Brady’s 266.
+While Brady’s rushing stats are modest, with 1,256 rushing yards and 35 rushing touchdowns, Mahomes is projected to amass 3,348 rushing yards and 40 rushing touchdowns, showing a clear advantage in rushing ability.
 This emphasizes Mahomes’ dual-threat capability, which contributes to his versatility as a quarterback.
@@
 The analysis of per-game statistics shows a clear edge for Patrick Mahomes over Tom Brady across nearly every metric.
-Mahomes completes slightly more passes per game (23.94) than Brady (23.50), despite attempting a comparable number of throws (36.74 for Mahomes vs.
+Mahomes completes slightly more passes per game (23.96) than Brady (23.50), despite attempting a comparable number of throws (36.81 for Mahomes vs.
 36.67 for Brady).
-This small difference results in a higher completion percentage for Mahomes at 65.17%, compared to Brady’s 64.08%, indicating a marginal but noteworthy advantage in passing efficiency.
-Mahomes also outpaces Brady in passing yards per game, averaging 276.38 to Brady’s 269.33, further emphasizing his ability to generate more offense through the air.
+This small difference results in a higher completion percentage (defined as the number of completions divided by the number of attempts) for Mahomes at 65.10%, compared to Brady’s 64.08%, indicating a marginal but noteworthy advantage in passing efficiency.
+Mahomes also outpaces Brady in passing yards per game, averaging 275.18 to Brady’s 269.33, further emphasizing his ability to generate more offense through the air.
 
-In terms of scoring, Mahomes demonstrates a superior passing touchdown rate, averaging 2.04 touchdowns per game compared to Brady’s 1.94.
-Both players maintain an impressive level of ball security, with nearly identical interception rates per game (0.66).
+In terms of scoring, Mahomes demonstrates a superior passing touchdown rate, averaging 2.02 touchdowns per game compared to Brady’s 1.94.
+Both players maintain an impressive level of ball security, with nearly identical interception rates per game (0.66 for Brady to 0.67 for Mahomes).
 
 The most significant disparity appears in rushing statistics, where Mahomes outshines Brady by a wide margin.
-Mahomes averages 10.13 rushing yards per game, more than triple Brady’s 3.30, and his higher number of carries per game (2.85 vs.
+Mahomes averages 10.33 rushing yards per game, more than triple Brady’s 3.30, and his higher number of carries per game (2.87 vs.
 2.12) underscores his ability to contribute as a dual-threat quarterback.
-Furthermore, Mahomes’ rushing touchdown rate (0.11 per game) surpasses Brady’s 0.09, further illustrating his dynamic impact on the ground.
+Furthermore, Mahomes’ rushing touchdown rate (0.12 per game) surpasses Brady’s 0.09, further illustrating his dynamic impact on the ground.
 

### 2024-12-01 models/model_1.R
update model

@@ -9,6 +9,7 @@
 # Load necessary libraries
 library(dplyr)
 library(arrow)
+library(caret)  # For the createDataPartition function to split data
 
 # Load the dataset
 file_path <- "~/nfl/data/analysis_data/game_stats/merged_player_game.parquet"
@@ -18,13 +19,10 @@ player_data <- read_parquet(file_path)
 brady_data <- player_data %>% filter(player_name == "T.Brady")
 mahomes_data <- player_data %>% filter(player_name == "P.Mahomes")
 
-# Select numeric columns for regression
+# Select relevant numeric columns for regression
 numeric_features <- c(
-  "completions", "attempts", "passing_yards", "passing_tds",
-  "interceptions", "sack_fumbles", "sack_fumbles_lost",
-  "passing_air_yards", "passing_first_downs", "passing_2pt_conversions",
-  "carries", "rushing_yards", "rushing_tds", "rushing_fumbles",
-  "rushing_fumbles_lost", "rushing_first_downs"
+  "completions", "attempts", "passing_yards", "passing_tds", 
+  "interceptions", "carries", "rushing_yards", "rushing_tds"
 )
 
 # Ensure no missing values in selected columns for Tom Brady
@@ -32,16 +30,39 @@ brady_data_clean <- brady_data %>%
   select(all_of(numeric_features)) %>%
   na.omit()
 
-# Create the linear regression model
+# Set seed for reproducibility
+set.seed(8)
+
+# Create a 80/20 train-test split
+train_index <- createDataPartition(brady_data_clean$completions, p = 0.80, list = FALSE)
+train_data <- brady_data_clean[train_index, ]
+test_data <- brady_data_clean[-train_index, ]
+
+# Create the linear regression model using Brady's training data
 reg_model <- lm(
   cbind(
-    completions, attempts, passing_yards, passing_tds, interceptions,
-    sack_fumbles, sack_fumbles_lost, passing_air_yards, passing_first_downs,
-    passing_2pt_conversions, carries, rushing_yards, rushing_tds,
-    rushing_fumbles, rushing_fumbles_lost, rushing_first_downs
+    completions, attempts, passing_yards, passing_tds, interceptions, 
+    carries, rushing_yards, rushing_tds
   ) ~ ., 
-  data = brady_data_clean
+  data = train_data
 )
+
+# Use the trained model to predict on Brady's test data
+predicted_brady_stats <- predict(
[…]

### 2024-12-01 scripts/clean_data.R
Update clean_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Cleans NFL data for Tom Brady and Patrick Mahomes
 # Author: Alexander Guarasci
-# Date: 26 November 2024
+# Date: 1 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `nflverse` and `arrow` packages must be installed, run download_data.R 

### 2024-12-01 scripts/download_data.R
Update download_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Downloads NFL data for Tom Brady and Patrick Mahomes
 # Author: Alexander Guarasci
-# Date: 26 November 2024
+# Date: 1 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `nflverse` and `arrow` packages must be installed

### 2024-12-01 scripts/simulate_game_data.R
Update preamble

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Simulates and saves a dataset containing NFL game data for Tom Brady and Patrick Mahomes
 # Author: Alexander Guarasci
-# Date: 26 November 2024
+# Date: 1 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `dplyr` and `arrow` packages must be installed.

### 2024-12-01 scripts/simulate_qbr_data.R
Update preamble

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Simulates and saves a dataset containing select QBR data for Tom Brady and Patrick Mahomes, retaining only specified columns.
 # Author: Alexander Guarasci
-# Date: 26 November 2024
+# Date: 1 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `dplyr` and `arrow` packages must be installed.

### 2024-12-01 scripts/test_analysis_qbr_data.R
Update preamble

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Tests analysis NFL QBR data for Tom Brady and Patrick Mahomes
 # Author: Alexander Guarasci
-# Date: 26 November 2024
+# Date: 1 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `dplyr`, `arrow` and 'testthat' packages must be installed, and clean_data.R must be run

### 2024-12-01 scripts/test_game_analysis_data.R
Update preamble

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Tests analysis NFL game data for Tom Brady and Patrick Mahomes
 # Author: Alexander Guarasci
-# Date: 26 November 2024
+# Date: 1 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `dplyr`, `arrow` and 'testthat' packages must be installed, and clean_data.R must be run

### 2024-12-01 scripts/test_simulated_game_data.R
Update preamble

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Tests simulated NFL game data for Tom Brady and Patrick Mahomes
 # Author: Alexander Guarasci
-# Date: 26 November 2024
+# Date: 1 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `dplyr`, `arrow` and 'testthat' packages must be installed, and simulate_game_data.R must be run

### 2024-12-01 scripts/test_simulated_qbr_data.R
Update test_simulated_qbr_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Tests simulated NFL QBR data for Tom Brady and Patrick Mahomes
 # Author: Alexander Guarasci
-# Date: 26 November 2024
+# Date: 1 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `dplyr`, `arrow` and 'testthat' packages must be installed, and simulate_qbr_data.R must be run

### 2024-12-01 models/model_1.R
Update preamble

@@ -1,10 +1,10 @@
 #### Preamble ####
 # Purpose: Predict Patrick Mahomes' lifetime career stats for 200 additional games
 # Author: Alexander Guarasci
-# Date: November 20, 2024
+# Date: 1 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: The `tidyverse` and `arrow` packages must be installed.
+# Pre-requisites: The `tidyverse`, `caret` and `arrow` packages must be installed.
 
 # Load necessary libraries
 library(dplyr)

### 2024-12-01 paper/paper.qmd
think final update for paper

@@
     fig-pos: "H"
-number-sections: false
+number-sections: true
 bibliography: references.bib
@@
 
-The paper is structured as follows: the first section provides an explanation of the data used in this paper, what it shows and where it comes from.
-The second section discusses the model used to predict the future of Patrick Mahomes’ career.
-The third section shows the results of the model.
-The final section discusses the implication of the model as well as provides more insight into the two players’ legacies.
+The paper is structured as follows: @sec-data provides an explanation of the data used in this paper, what it shows and where it comes from.
+@sec-model discusses the model used to predict the future of Patrick Mahomes’ career.
+@sec-results shows the results of the model.
+@sec-discussion discusses the implication of the model as well as provides more insight into the two players’ legacies.
 
@@
 The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Caret [@citecaret], Arrow [@citearrow], Knitr [@citeknitr], and Testthat [@citett] packages.
-The primary dataset included QB-specific data from every NFL game played since 2000, encompassing over 55 columns detailing various player performance metrics.
+The code was styled using [@citestyler] The primary dataset included QB-specific data from every NFL game played since 2000, encompassing over 55 columns detailing various player performance metrics.
 Additional data sources were used for ESPN’s Quarterback Rating (QBR), which encapsulates QB efficiency across multiple dimensions of gameplay.
@@
   geom_bar(stat = "identity") +
-  scale_fill_manual(values = c("red","blue")) +
+  scale_fill_manual(values = c("red", "blue")) +
   theme_minimal() +
@@
     y = "Passing Yards",
-    fill = ''
+    fill = ""
   )
-
 ```
@@
     y = "Rushing Yards",
-    fill = ''
+    fill = ""
   )
@@
   )
-
 ```
@@
   ggplot(aes(x = player_display_name, y = passing_yards, fill = player_display_name)) +
-  geom_violin(alpha = 0.7,trim = FALSE) +
+  geom_violin(alpha = 0.7, trim = FALSE) +
   theme_minimal() +
@@
     y = "QBR",
-    fill = 'Player'
+    fill = "Player"
   )
@@
 
-# Model
+# Model {#sec-model}
 
@@
 - **attempts**: The total number of pass attempts.
-- **passing_yards**: The total yards gained from passing attempts.
-- **passing_tds**: The total number of passing touchdowns.
+- **passing yards**: The total yards gained from passing attempts.
+- **passing tds**: The total number of passing touchdowns.
 - **interceptions**: The total number of interceptions thrown.
 - **carries**: The number of rushing attempts.
-- **rushing_yards**: The total rushing yards gained.
-- **rushing_tds**: The total number of rushing touchdowns.
+- **rushing yards**: The total rushing yards gained.
+- **rushing tds**: The total number of rushing touchdowns.
 
-The model used in this analysis is a multiple linear regression model.
+The model used in this analysis is a multiple linear regression model (@eq-lm).
 We choose linear regression because it is a simple yet effective approach for modeling continuous outcomes and provides easy interpretability.
@@
 Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_n X_n + \epsilon
-$$
+$$ {#eq-lm}
 
@@
 We also calculated the Root Mean Squared Error (RMSE) for each predicted variable in the test set as an evaluation metric.
-RMSE is a commonly used metric to measure the accuracy of regression models, and it is calculated as:
+RMSE (@eq-rmse) is a commonly used metric to measure the accuracy of regression models, and it is calculated as:
 
@@
 RMSE = \sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2}
-$$
+$$ {#eq-rmse}
 
@@
 
-## Limitiations
+## Assumptions and Limitations
 
-While the RMSE values suggest some error in the predictions, this is expected due to the limited amount of available data.
-The performance of the model could likely improve with a larger and more balanced dataset that includes a broader range of game conditions.
-Additionally, while we used Tom Brady’s statistics as a proxy for Mahomes, differences in their playing styles and team dynamics could also contribute to the prediction error.
+The model makes several key assumptions that are important for interpreting the predictions:
 
-The linear regression model assumes that the relationships between the predictors and the outcomes are linear, which may not always be the case in real-world scenarios.
-Further exploration of alternative models, such as non-linear regression or machine learning techniques like decision trees, could be considered for improving predictions.
+1.
+**Similarity of Player Performance**: The model assumes that Patrick Mahomes' future performance will be similar to Tom Brady’s past performance, using Brady's career as a proxy.
+This assumption may not hold as Mahomes' playing style and team dynamics significantly differ from Brady's.
+
+2.
+**Linear Relationships**: The model assumes that the relationships between the predictors (game statistics such as completions, passing yards, etc.) and the outcomes are linear.
+This may not always be the case in practice, as player performance could be influenced by more complex, non-linear interactions.
+
+3.
+**Constant Game Conditions**: The model assumes that game conditions, including factors such as opposing defenses and weather conditions, remain relatively consistent over time.
+However, these factors could vary and influence player performance in ways not captured by the model.
+
+4.
+**Prediction for 200 Games**: The model assumes that Mahomes' performance will follow a similar trajectory over the next 200 games as it has in the past.
+This assumption may not account for potential career milestones, injuries, or other events that could disrupt this pattern.
+
+5.
+**Uniform Dataset**: The model assumes that the dataset of Tom Brady’s career is representative of typical game conditions and does not contain significant biases or outliers that would skew the predictions.
+
+Ultimately, the model may be appropriate in scenarios where general trends in quarterback performance are needed for broad projections, such as for media analysis or high-level discussions of player comparisons.
+It can offer insights into how Mahomes might perform in a typical game based on historical data from a similar elite quarterback.
+This can be useful for estimating the overall trajectory of Mahomes’ career and projecting his future career stats across a large sample of games.
+
+However, the model may not be appropriate for precise, game-by-game predictions, especially when specific game conditions or player-specific factors (such as injury status or team dynamics) are likely to vary.
+For example, using this model to predict Mahomes’ performance in a critical playoff game or during a season affected by significant roster changes could lead to inaccurate or misleading conclusions.
+Additionally, the linear nature of the model means it may fail to account for more complex interactions that influence a quarterback’s performance in real-world scenarios.
+
+In summary, while the model provides a reasonable estimation of Mahomes’ future statistics based on Brady’s historical data, it should be used cautiously in contexts where precision is critical or where Mahomes’ unique playing style and career circumstances may diverge from Brady’s past performance.
 
@@
 
-# Results
+# Results {#sec-results}
  [@tbl-brady-mahomes-career-stats] compares Tom Brady’s career statistics with Patrick Mahomes’ projected lifetime statistics.
@@
 # Generate the table using tinytable
-tt(final_table, width = 1.5) |> style_tt( fontsize = 1)
+tt(final_table, width = 1.5) |> style_tt(fontsize = 1)
 ```
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 

### 2024-12-01 paper/paper.qmd
Add files via upload

@@
 The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Caret [@citecaret], Arrow [@citearrow], Knitr [@citeknitr], and Testthat [@citett] packages.
-The code was styled using [@citestyler] The primary dataset included QB-specific data from every NFL game played since 2000, encompassing over 55 columns detailing various player performance metrics.
+The code was styled using Styler [@citestyler].
+The primary dataset included QB-specific data from every NFL game played since 2000, encompassing over 55 columns detailing various player performance metrics.
 Additional data sources were used for ESPN’s Quarterback Rating (QBR), which encapsulates QB efficiency across multiple dimensions of gameplay.

### 2024-12-03 paper/paper.qmd
add reference

@@
 
-The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Caret [@citecaret], Arrow [@citearrow], Knitr [@citeknitr], and Testthat [@citett] packages.
+The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Caret [@citecaret], Arrow [@citearrow],Tinytable [@citetinytable], Knitr [@citeknitr], and Testthat [@citett] packages.
 The code was styled using Styler [@citestyler].

### 2024-12-03 paper/paper.qmd
fix spelling error

--- 2024-12-01 paper/paper.qmd
@@
 
-The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Caret [@citecaret], Arrow [@citearrow], Knitr [@citeknitr], and Testthat [@citett] packages.
+The analysis was conducted in R [@citeR], leveraging tools from the Tidyverse [@citetidyverse], Caret [@citecaret], Arrow [@citearrow], Tinytable [@citetinytable], Knitr [@citeknitr], and Testthat [@citett] packages.
 The code was styled using Styler [@citestyler].

### 2024-12-03 scripts/clean_data.R
Update date

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Cleans NFL data for Tom Brady and Patrick Mahomes
 # Author: Alexander Guarasci
-# Date: 1 December, 2024
+# Date: 3 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `nflverse` and `arrow` packages must be installed, run download_data.R 

### 2024-12-03 scripts/download_data.R
Update date

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Downloads NFL data for Tom Brady and Patrick Mahomes
 # Author: Alexander Guarasci
-# Date: 1 December, 2024
+# Date: 3 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `nflverse` and `arrow` packages must be installed

### 2024-12-03 scripts/simulate_game_data.R
Update date

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Simulates and saves a dataset containing NFL game data for Tom Brady and Patrick Mahomes
 # Author: Alexander Guarasci
-# Date: 1 December, 2024
+# Date: 3 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `dplyr` and `arrow` packages must be installed.

### 2024-12-03 scripts/simulate_qbr_data.R
Update date

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Simulates and saves a dataset containing select QBR data for Tom Brady and Patrick Mahomes, retaining only specified columns.
 # Author: Alexander Guarasci
-# Date: 1 December, 2024
+# Date: 3 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `dplyr` and `arrow` packages must be installed.

### 2024-12-03 scripts/test_analysis_qbr_data.R
Update date

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Tests analysis NFL QBR data for Tom Brady and Patrick Mahomes
 # Author: Alexander Guarasci
-# Date: 1 December, 2024
+# Date: 3 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `dplyr`, `arrow` and 'testthat' packages must be installed, and clean_data.R must be run

### 2024-12-03 scripts/test_game_analysis_data.R
Update date

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Tests analysis NFL game data for Tom Brady and Patrick Mahomes
 # Author: Alexander Guarasci
-# Date: 1 December, 2024
+# Date: 3 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `dplyr`, `arrow` and 'testthat' packages must be installed, and clean_data.R must be run

### 2024-12-03 scripts/test_simulated_game_data.R
Update date

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Tests simulated NFL game data for Tom Brady and Patrick Mahomes
 # Author: Alexander Guarasci
-# Date: 1 December, 2024
+# Date: 3 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `dplyr`, `arrow` and 'testthat' packages must be installed, and simulate_game_data.R must be run

### 2024-12-03 scripts/test_simulated_qbr_data.R
Update date

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Tests simulated NFL QBR data for Tom Brady and Patrick Mahomes
 # Author: Alexander Guarasci
-# Date: 1 December, 2024
+# Date: 3 December, 2024
 # Contact: alexander.guarasci@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: The `dplyr`, `arrow` and 'testthat' packages must be installed, and simulate_qbr_data.R must be run

### 2024-12-03 models/model_1.R
Update model_1.R

@@ -1,11 +1,3 @@
-#### Preamble ####
-# Purpose: Predict Patrick Mahomes' lifetime career stats for 200 additional games
-# Author: Alexander Guarasci
-# Date: 1 December, 2024
-# Contact: alexander.guarasci@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: The `tidyverse`, `caret` and `arrow` packages must be installed.
-
 # Load necessary libraries
 library(dplyr)
 library(arrow)
